### 2019-04-02 manuscript/manuscript.Rmd
Starting framework for manuscript

---
title: "Lake Photic Zone Temperature accross the Conterminous United States"
author:
- Kreakie, B. J. ^\*^ ^*1*^, Milstead, W. Bryan. ^*1*^, Hollister. J. W. ^*1*^, Shivers,
  S. ^*2*^, 
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'

bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: inline
csl: plos.csl

abstract: Text for abstract

---

```{r setup, include=FALSE}
source(here::here("R/functions.R"), encoding = "UTF-8")
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
options(scipen = 2, digits = 5)
```

```{r analysis, include=FALSE}
#got to figure out path stuff before render.
source(here::here("R/data_available.R"), encoding = "UTF-8") 
source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
source(here::here("R/figures.R"), encoding = "UTF-8")
```

# Introduction

# Methods

# Results

# Discussion and conclusions

# Bibliography

### 2019-04-02 manuscript/manuscript.Rmd
small changes

@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, eval=FALSE}
 source(here::here("R/functions.R"), encoding = "UTF-8")
@@
 
-```{r analysis, include=FALSE}
+```{r analysis, include=FALSE, eval=FALSE}
 #got to figure out path stuff before render.

### 2019-04-02 R/file.R
playing

@@ -1,0 +1,1 @@
+text and code

### 2019-04-03 manuscript/manuscript.Rmd
Added text to the intro

@@
 
+During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world (Williamson et al.
+2009, Schindler 2009).
+As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)(NASA GISS), Surface temperatures of lakes are also increasing globally (0.34 per decade from 1985 to 2009) [@o2015rapid].
+The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.
+Ultimately, temperature changes will greatly impact all aspects of lake resource management.
+For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].
+During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].
+As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over other phytoplankton and can proliferate quickly [@paerl2008blooms].
+Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].
+Thus, our ability to understand and predict toxin cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.
+This need for accurate lake temperature forecasting is will become crucial for protecting human and environmental health.
+
+Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.
+These studies typically vary in number of lakes studied, complexity of modelling approach, and time interval.
+Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).
+Such studies include: [@saeed2016water; @peeters2002modelin; @zhong2016recent].
+There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
+By and large, air temperature and lake size are often the only selected predictor variables.
+Finally, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.
+Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].
+Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.
+Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes (Downing et al.
+2006, Winslow et al.
+2014).
+
 # Methods

### 2019-04-26 R/model.Rmd
Creating initial RF run with all data

---
title: "Untitled"
author: "Betty Kreakie"
date: "April 25, 2019"
output: html_document
---

```{r libraries, echo=TRUE, cache=TRUE}

library(randomForest)
library(dplyr)
library(purrr)
library(here)
library(broom)
library(ggplot2)
```

#Run Variable Selection stuff (hollister et al 2017)

```{r var_sel,echo=TRUE, cache=TRUE}

#varsel
source(here::here('R/varsel_regression_rf.R'))
source(here::here('R/varsel_plot.R'))

x<-complete.cases(nla)
nla.cc<-nla[x,]
nla.cc$year<-factor(nla.cc$year)

data07 <- filter(nla.cc, year == 2007, tmean_2m < 70) %>%
            dplyr::select(tmean_2m, date, longitude, latitude, tmean, tmean_dm1,
            tmean_avg3, tmean_avg7, tmean_avg30, percent_impervious, surface_area,
            shoreline_length, shoreline_dev, max_length, max_width, mean_width,
            max_depth, max_depth, volume, elevation) # 1214   19

data12 <- filter(nla.cc, year == 2012, tmean_2m < 70) %>%
            dplyr::select(tmean_2m, date, longitude, latitude, tmean, tmean_dm1,
            tmean_avg3, tmean_avg7, tmean_avg30, percent_impervious, surface_area,
            shoreline_length, shoreline_dev, max_length, max_width, mean_width,
            max_depth, max_depth, volume, elevation)

nla.all <-  select(nla.cc, tmean_2m, year, date, longitude, latitude, tmean,
                   tmean_dm1, tmean_avg3, tmean_avg7, tmean_avg30,percent_impervious,                    surface_area,shoreline_length, shoreline_dev, max_length,                            max_width, mean_width, max_depth, max_depth, volume, elevation)

var_sel_tmean07 <- varsel_regression_rf(data07$tmean_2m, select(data07, -tmean_2m),                ntree=1000, prog=TRUE)

varsel_plot(var_sel_tmean07) #9 variables

var_sel_tmean07$vars[[9]]

var_sel_tmean12 <- varsel_regression_rf(data12$tmean_2m, select(data12, -tmean_2m),                ntree=1000, prog=TRUE)

varsel_plot(var_sel_tmean12) #9 variables

var_sel_tmean12$vars[[9]]

var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE)

varsel_plot(var_sel_tmean) #9 variables

var_sel_tmean$vars[[9]]

```

# Build our draft model with selected variables

```{r draft_model,echo=TRUE, cache=TRUE}

# Draft Model

# Using tmean made most sense and perfomed better that tmin and tmax based models.

RF2007 <- data07 %>%

  select(tmean_2m,var_sel_tmean07$vars[[9]]) %>%

  randomForest(tmean_2m ~ ., data = ., importance = TRUE,

               ntree = 10000)

RF2007

varImpPlot(RF2007)

RF2012 <- data12 %>%

  select(tmean_2m,var_sel_tmean12$vars[[9]]) %>%

  randomForest(tmean_2m ~ ., data = ., importance = TRUE,

               ntree = 10000)

RF2012

varImpPlot(RF2012)

RFAll <- nla.all %>%

  select(tmean_2m,var_sel_tmean$vars[[9]]) %>%

  randomForest(tmean_2m ~ ., data = ., importance = TRUE,

               ntree = 10000)

RFAll

varImpPlot(RFAll)

```

mean_run <- nla.cc %>%

  select(tmean_2m,var_sel_tmean$vars[[9]]) %>%

  randomForest(tmean_2m ~ ., data = ., importance = TRUE,

               ntree = 1000)

mean_run

varImpPlot(mean_run)

```

### 2019-04-26 data/lake_temperatures_data.Rmd
Creating initial RF run with all data

---
title: "lake_temperatures_data"
author: "bryan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(prism)
library(readr)
library(raster)
library(mapview)
library(knitr)
# library(goatscape) #use source(here::here('R/gs_nlcd.r')) instead
library(prism)
library(elevatr)
# library(goatscape)  # use: source(here::here('R/gs_nlcd.r'))
library(RSQLite)
library(readxl)

options(stringsAsFactors=FALSE)
options(prism.path = "L:/Public/Milstead_Lakes/prism")  # will be really big

# create dir "data_local" for scratch files
if(!dir.exists(here::here('data_local'))) dir.create(here::here('data_local'))

# load functions
source(here::here('R/gs_nlcd.r'))  # the goatscape gs_nlcd.r function

```

## To Do List
*

## Introduction
* The goal of this project is to compile all of the data necessary to model lake surface temperatures and make predictions for all lakes in NHDplus 
* This file replaces two preliminary attempts:
    * lake_temp_data.Rmd
    * prism_big_data.Rmd
* All data will be assembled and stored as a SQLite database (lake_temperatures.sqlite) 
    * database too large to share on github.  Must be assembled with the code below.
    * [SQLite tables and fields schema and data definitions] (https://github.com/willbmisled/photic_zone_temp/blob/master/data/lake_temperatures_sqlite.xlsx)
* This document also makes a subset of the data available as a .csv file.  This file [nla_base.csv] includes just the data for NLA lakes (2007 & 2012) that will be used to build a random forest model to predict lake temperatures.  Data definitions are at the bottom of this document.

## Data sources

* **lmorpho**: [Jeff Hollister's Lake Morphometry database](https://edg.epa.gov/data/PUBLIC/ORD/NHEERL/LakeMorphometry.zip)
* **nla**
    - [NLA2007 Lake locations shapefile](https://www.epa.gov/sites/production/files/2017-10/national_lakepoly_withmetrics_20090929.zip)
    - [NLA2012 Lake locations shapefile](https://www.epa.gov/sites/production/files/2017-10/nla_2012_sites_sampled.zip)
    - [NLA2007 design file](https://www.epa.gov/sites/production/files/2014-01/nla2007_sampledlakeinformation_20091113.csv"))
    - [NLA2012 design file ](https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_siteinfo_08232016.csv)
    - [NLA2007 profile data with lake temperatures by depth](https://www.epa.gov/sites/production/files/2013-09/nla2007_profile_20091008.csv) 
    - [NLA2012 profile data with lake temperatures by depth](https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_profile_08232016.csv)
* **nlcd**: [2006 & 2011 NLCD percent impervious cover in a 3000m buffer around lake](https://www.mrlc.gov/index.php)
* **prism**: [prism estimates of daily min & max temperatures for date range (sample_date - 30 days):(sample_date) by location](http://www.prism.oregonstate.edu/)

## CRS info
* prism: '+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0'
* centroids: '+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0'
* lmorhpho: "+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext +no_defs"
* nlcd: "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0
+units=m +no_defs"

## Data steps

* Download the tmean prism daily for May1-Sep 30 for years 1981:2018

```{r get_prism, include=FALSE, echo=FALSE, eval= FALSE}
# NOte this will take many hours

for(i in 1981:2018) {
# for(i in c(1981:2006, 2008:2011, 2013:2018)) {
  min <- ymd(paste0(i, "05", "01"))
  max <- ymd(paste0(i, "09", "30"))
  print(paste(i, Sys.time()))
  get_prism_dailys(type="tmean", minDate = min, maxDate = max, keepZip=FALSE)
}}

```

* Lake Morphometry
    - download, convert to a single sf file
    - save as "lmorpho.rda"

```{r get_morpho, include=FALSE, echo=FALSE, eval=FALSE} 
#download existing lake morphometry
if(!dir.exists(here::here("data_local/LakeMorphGdb.gdb"))){
  temp <- tempfile()
  download.file("https://edg.epa.gov/data/PUBLIC/ORD/NHEERL/LakeMorphometry.zip",temp)
  unzip(temp, exdir = here::here('data_local'))
  unlink(temp)
}

# get layers
if(file.exists(here::here('data_local/lmorpho.rda'))) {
  load(here::here('data_local/lmorpho.rda'))
  } else {
    lyrs <- sf::st_layers(here::here("data_local/LakeMorphGdb.gdb"))
    lmorph <- purrr::map(lyrs$name, function(x) sf::read_sf(here::here("data_local/LakeMorphGdb.gdb/"), x))
    lmorpho <- do.call(rbind, lmorph)

  # change nlaSITE_ID == "NA" to NA
  dim(lmorpho) # 363314     19
  table(is.na(lmorpho$nlaSITE_ID)) # FALSE 363314

  lmorpho <- mutate(lmorpho, nlaSITE_ID = ifelse(nlaSITE_ID == "NA", NA, nlaSITE_ID))
  table(is.na(lmorpho$nlaSITE_ID)) # FALSE =1151  TRUE = 362163

  save(lmorpho, file = here::here('data_local/lmorpho.rda'))
}
```

* add centroids (longitude and latitude) to lmorpho
    * extract centroids from lmorpho
    * transform to prism crs (+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0)
    * add as longitude and latitude

```{r centroids, include=FALSE, echo=FALSE, eval=FALSE} 
# download the lmorpho data
load(here::here('data_local/lmorpho.rda'))

# get the centroids
centroids <- st_centroid(lmorpho)

# transform to prism CRS
# +proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0

centroids <- st_transform(centroids, "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0")

# add lat long and convert to data.frame
xy <- as.data.frame(st_coordinates(centroids))
names(xy) <- c('longitude', 'latitude')

centroids <- as.data.frame(centroids) %>%
    dplyr::select(-Shape) %>%
    cbind(xy) 
```

* get the prism cellnums (the grid cells) for the centroids

```{r cellnums, include=FALSE, echo=FALSE, eval=FALSE} 
# get a prism raster 
pf <- raster(ls_prism_data(absPath=T)[1, 2])

# change values of grid cells to the cell number
values(pf) <- 1:ncell(pf)

# extract the cellnums from the rasters
a <- extract(x = pf, y = centroids[ , c("longitude", "latitude")])
    length(a) # 363314
    length(unique(a)) # 144402

# add cellnum to centroids
centroids$cellnum <- a

# save file
save(centroids, file = here::here('data/centroids.rda'))
```

* write a function to extract the prism data
    * each year will be separate file saved as data_local/meanYYYY.rds  with YYYY = year

```{r func_prism, include=FALSE, echo=FALSE, eval=TRUE}
# function to extract the values for all dates in a year
pvals <- function(year) {
# free up memory
  gc()
# get the dates
  dates <- format(seq(ymd(paste0(year, "05", "01")), ymd(paste0(year, "09", "30")), by="days"), format="%Y%m%d")
# get the prism file names
  pfiles <- paste0('PRISM_tmean_stable_4kmD1_', dates, '_bil')
# make a raster stack
  pstack <- prism_stack(pfiles)
# get the values for all cellnums
  vals <- as.data.frame(getValues(pstack))
# shorten names to dmmdd (e.g., d0501)
  names(vals) <- paste0("d", substr(dates, 5, 8))
# add cellnum (same as the rowname)
  vals$cellnum <- as.numeric(rownames(vals)) 
# add year
  vals$year <- year
# keep only cellnums (cnums) with lakes in them
  vals <- left_join(cnums, vals) 
# assign df names as meanYYYY (year from input)
  fname <- paste0('mean', year)
  assign(fname, vals)
# save the file
  outname <- paste0('data_local/', fname, '.rds')
  saveRDS(object = get(fname), file = here::here(outname)) 
}

```

* Extract the values from the prism rasters and add to the table tmeans_raw in lake_temperatures.sqlite

```{r vals_prism, include=FALSE, echo=FALSE, eval=FALSE}
# create a list of cellnums is available
if(exists("cnums") == FALSE){
  if(exists("centroids") == FALSE) load(here::here('data/centroids.rda'))
  cnums <- data.frame(cellnum = unique(centroids$cellnum)) %>% # 144402      1
    arrange(cellnum)
}

# loop to extract values (note a file for each year will be saved to /data/ but not stored in memory)
for (i in c(1981:2017)) {
  if(!file.exists(paste0('data_local/mean', i, '.rds'))) {
    print(i)
    print(Sys.time())
    pvals(i)
}}

# combine all means into a single SQLite table

# create or connect to the database
  means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

# append data to table means
  for(i in c(1981:2017)) {
  print(Sys.time())
  print(i)
  rds <- paste0('data_local/mean', i, '.rds')
  a <- readRDS(here::here(rds))
  dbWriteTable(means, "means", a, append = TRUE)
  print(memory.size())
  }

# let's rename table 'means' to 'tmeans_raw'
dbListTables(means)
dbExecute(means, 'ALTER TABLE means RENAME TO tmeans_raw')
dbListFields(means, "tmeans_raw")

# create indices for cellnum and year for table "tmeans_raw"
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
dbExecute(means,"CREATE INDEX index_tmeans_raw_cnum ON tmeans_raw (cellnum)")
dbExecute(means,"CREATE INDEX index_tmeans_raw_year ON tmeans_raw (year)")
dbDisconnect(means)

# close and save the database
  dbDisconnect(means)
```

* calculate 3, 7, and 30 mean temps
    * means calculated and added to the table tmeans in lake_temperatures.sqlite
    * this takes a long time to eval = FALSE; change if you want to rerun

```{r means_prism, include=FALSE, echo=FALSE, eval=FALSE}

for(j in c(1983:2017)) { 
# for(j in c(1981:1982)) {   
  print(Sys.time())
# load a year of data
  start <- Sys.time()
  print(j)
  rds <- paste0('data_local/mean', j, '.rds')
  tmean <- readRDS(here::here(rds))

# create dfs to hold the preliminary results
  tmean3 <- tmean[ , -c(2:32)] 
  tmean3[ , 2:123] <- NA
  tmean7 <- tmean3
  tmean30 <- tmean3

# create df of temps for day-1
  tmean_dm1 <- tmean3
  tmean_dm1[ ,2:123] <- tmean[ ,32:153]

# loop to calculate the 3, 7, and 30 day means 
for(i in c(33:154)){
  tmean3[ , i-31] <- apply(tmean[ ,(i-3):(i-1)], 1, mean)
  tmean7[ , i-31] <- apply(tmean[ ,(i-7):(i-1)], 1, mean)
  tmean30[ , i-31] <- apply(tmean[ ,(i-30):(i-1)], 1, mean)
}

# convert to long format
  tm <- gather(tmean[ , -c(2:32)], date, tmean, -cellnum, -year)
  tm_dm1 <- gather(tmean_dm1, date, tmean_dm1, -cellnum, -year)
  tm3 <- gather(tmean3, date, tmean, -cellnum, -year)
  tm7 <- gather(tmean7, date, tmean, -cellnum, -year)
  tm30 <- gather(tmean30, date, tmean, -cellnum, -year)

# combine year and date
  tm <- dplyr::mutate(tm, date = paste0(tm$year, substr(tm$date, 2, 5))) %>%
    dplyr::select(-year)

# add dm1 and 3, 7, and 30 day means
  tm$tmean_dm1 <- tm_dm1$tmean_dm1
  tm$tmean_avg3 <- tm3$tmean
  tm$tmean_avg7 <- tm7$tmean
  tm$tmean_avg30 <- tm30$tmean

# connect to the database
  means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

# append data to table tmeans
  dbWriteTable(means, "tmeans", tm, append = TRUE)

# QAQC
  # dbListTables(means)
  # dbGetQuery(means, 'SELECT Count(*) FROM tmeans') # 17617044 (after first upload)
  # dim(tm) # 17617044        6

# close and save the database
  dbDisconnect(means)

print(Sys.time() - start)
}

# create indices for cellnum and date for table "tmeans"
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
dbExecute(means,"CREATE INDEX index_tmeans_cnum ON tmeans (cellnum)") 
dbExecute(means,"CREATE INDEX index_tmeans_date ON tmeans (date)")
dbDisconnect(means)

# QAQC
# check number of lines
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
dbGetQuery(means, 'SELECT Count(*) FROM tmeans') # 17617044 * 37 = 651,830,628

# check some of the means for a random cellnums and dates
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

yr1 <- sample(1981:2017, 1)
day1 <- sample(format(seq(ymd(paste0(yr1, "06", "01")), ymd(paste0(yr1, "09", "30")), by="days"), format="%m%d"), 1)
date1 <- paste0(yr1, day1)

mean1 <- dbGetQuery(means, paste0('SELECT * FROM tmeans WHERE cellnum = ', cn, ' AND date = ', date1))

mean2 <- dbGetQuery(means, paste0('SELECT * FROM tmeans_raw WHERE cellnum = ', cn, ' AND year = ', yr1))

dbDisconnect(means)

i <- which(names(mean2) == paste0('d', day1))
c(mean1$tmean == mean2[i],
mean1$avg_3day_tmean == apply(mean2[ ,(i-2):i], 1, mean),
mean1$avg_7day_tmean == apply(mean2[ ,(i-6):i], 1, mean),
mean1$avg_30day_tmean == apply(mean2[ ,(i-29):i], 1, mean))
##

#

# qa qc repeat in excel for one year (use a subset of the cellnums) - results are the same
  # sample 100 cellnums from mean1981
mean2001 <- readRDS(here::here('data_local/mean2001.rds'))
temp <- mean2001[sample(as.numeric(row.names(mean2001)), 100), ] %>%
      arrange(cellnum)

  #get the calculated means for the sampled cellnums
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

aves <- dbGetQuery(means, paste0('SELECT * FROM tmeans WHERE date IN (20010601, 20010707, 20010930) AND
                                 cellnum IN (', paste(temp$cellnum, collapse=", "), ')')) %>%
        arrange(cellnum)
dbDisconnect(means)

  # write data to csv
write.csv(temp, here::here('data_local/temp/tmean.csv'))
write.csv(aves, here::here('data_local/temp/tmean_avg.csv'))

  # calculate the means in excel and compare to the results above
excel <- read.csv(here::here('data_local/temp/tmean_excel.csv'))
all.equal(aves$cellnum, excel$cellnum) # TRUE
all.equal(as.numeric(aves$date), excel$date) # TRUE
all.equal(aves$tmean, excel$tmean) # TRUE
all.equal(aves$tmean_dm1, excel$tmean_dm1) # TRUE
all.equal(aves$tmean_avg3, excel$tmean_avg3) # TRUE
all.equal(aves$tmean_avg7, excel$tmean_avg7) # TRUE
all.equal(aves$tmean_avg30, excel$tmean_avg30) # TRUE

```

* get the elevations for the centroids

```{r elevatr, include=FALSE, echo=FALSE, eval=FALSE}
locations <- dplyr::select(centroids, longitude, latitude, COMID)

prj <- "+proj=longlat +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +no_defs"
start <- Sys.time()
elev <- get_elev_point(locations, prj = prj, src = 'aws', z = 8)
end <- Sys.time()
end - start # 36.65486 mins

dim(centroids) # 363314     21
dim(elev) # 363314      3
centroids <- left_join(centroids, elev@data)
dim(centroids) # 363328     23

a <- as.data.frame(table(elev$COMID))
b <- dplyr::filter(a, Freq > 1)

dplyr::filter(elev@data, COMID %in% b$Var1)

# Why are there 14 extra points now?
length(unique(centroids$COMID)) # 363307  This includes the seve

saveRDS(object = elev, file = here::here('data/elevations.Rds'))

```

* create the lakes table from lmorpho, centroids, and elevation
* seven COMIDs removed because they were repeated and had slighly different data
* select and rename fields

```{r lakes, include=FALSE, echo=FALSE, eval=FALSE}
# check centroids / morpho for repeated comids
a <- as.data.frame(table(centroids$COMID))
b <- dplyr::filter(a, Freq > 1)  # 7 comids are repeated

lakes <- dplyr::filter(centroids, !COMID %in% b$Var1)
dim(centroids) # 363314     21
dim(lakes) # 363300     21; removed 7 centroids repeated 2 x each = 14 rows

# add elevation data
lakes <- left_join(lakes, elev@data) # 363300     23

# select and rename
lakes <- dplyr::select(lakes, cellnum, comid = COMID, longitude, latitude, 
                        surface_area = SurfaceArea,
                        shoreline_length = ShorelineLength,
                        shoreline_dev = ShorelineDev,
                        max_length = MaxLength, 
                        max_width = MaxWidth,
                        mean_width = MeanWidth,
                        max_depth = MaxDepthCorrect,
                        mean_depth = MeanDepthCorrect,
                        volume = VolumeCorrect,
                        elevation)

# create or connect to the database
  means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

# add lakes to table lakes
  dbWriteTable(means, "lakes", lakes, overwrite = TRUE)

# index table lakes
  dbExecute(means,"CREATE INDEX index_lakes_cnum ON lakes (cellnum)") 
  dbExecute(means,"CREATE INDEX index_lakes_comid ON lakes (comid)")

# housekeeping  
dbListTables(means)
dbListFields(means, "lakes")

# close and save the database
  dbDisconnect(means)

# save an Rds copy of lakes
  saveRDS(object = lakes, file = here::here('data/elevations.Rds'))

```

* NLA 
    - download the nla2007 & nla2012 shapefiles
    - transform to st_crs(lmorpho) 
        -  EPSG: 3857
        -  proj4string: "+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext +no_defs"
    - use st_join to match the 2007 and 2012 lakes to lmorpho
      - 54 nla07 lakes removed due to lack of 1:1 correspondence with lmorpho
      - 7 nla12 lakes removed because they did not map to any lmorpho lakes
      - 82 nla12 lakes removed due to lack of 1:1 correspondence with lmorpho
    - rbind nla2007 and nla2012

```{r get_nla_gis, include=FALSE, echo=FALSE, eval=FALSE}

# download nla07
  if(!file.exists(here::here('data_local/National_LakePoly_withMetrics_20090929.shp'))) {
    temp <- tempfile()
    download.file("https://www.epa.gov/sites/production/files/2017-10/national_lakepoly_withmetrics_20090929.zip",temp)
    unzip(temp, exdir = here::here('data_local'))
    unlink(temp)
  }

# download nla12
  if(!file.exists(here::here('data_local/NLA2012_Sampled_Polys_20131217.shp'))) {
    temp <- tempfile()
    download.file("https://www.epa.gov/sites/production/files/2017-10/nla_2012_sites_sampled.zip",temp)
    unzip(temp, exdir = here::here('data_local'))
    unlink(temp)
  }

# transform NLA shapefiles to st_crs(lmorpho)
  nla07 <- st_read(here::here('data_local/National_LakePoly_withMetrics_20090929.shp')) %>%
            st_transform(st_crs(lmorpho))  # dim = 1159    6

  nla12 <- st_read(here::here('data_local/NLA2012_Sampled_Polys_20131217.shp')) %>%
            st_transform(st_crs(lmorpho))  # dim = 1269 153

# join nla to lmorpho
  j07 <- st_join(nla07, lmorpho) %>%
          dplyr::select(nla_id = SITEID, comid = COMID) %>%
          mutate(year = 2007)

  j12 <- st_join(nla12, lmorpho) %>%
          dplyr::select(nla_id = NLA12_ID, comid = COMID.y) %>%
          mutate(year = 2012)

# convert joins to data.frames
  st_geometry(j07) <- NULL
  st_geometry(j12) <- NULL

# remove repeated comid's and and nla_ids; lakes without a 1:1 correspondence
#2007
  dim(nla07) # 1159    6
  dim(j07) # 1186   3
  table(is.na(j07$comid)) # FALSE = 1186
  table(is.na(j07$nla_id)) # FALSE = 1186

  a <- as.data.frame(table(j07$comid))
  bc <- filter(a, Freq > 1) # 4 lmorpho lakes map to multiple nla lakes map; delete these

  a <- as.data.frame(table(j07$nla_id))
  bs <- filter(a, Freq > 1) # 18 nla lakes map to multiple lmorpho lakes map; delete these

  j07 <- filter(j07, !comid %in% bc$Var1, !nla_id %in% bs$Var1) # 1132    3

#2012
  dim(nla12) # 1269  153
  dim(j12) # 1315    3
  table(is.na(j12$comid)) # FALSE = 1308 TRUE = 7
  table(is.na(j12$nla_id)) # FALSE = 1315

  j12 <- j12[!is.na(j12$comid), ] # 1308    3

  a <- as.data.frame(table(j12$comid))
  bc <- filter(a, Freq > 1) # 2 lmorpho lakes map to multiple nla lakes map; delete these

  a <- as.data.frame(table(j12$nla_id))
  bs <- filter(a, Freq > 1) # 32 nla lakes map to multiple lmorpho lakes map; delete these

  j12 <- filter(j12, !comid %in% bc$Var1, !nla_id %in% bs$Var1) # 1226    3

# rbind
  nla_id <- rbind(j07, j12) # 2358    3

# save an Rds copy of lakes
  saveRDS(object = nla_id, file = here::here('data/nla_id.Rds')) 
  # nla_id <- readRDS(file = here::here('data/nla_id.Rds'))
```

* get the nla profile data
* calc the mean temperature for top 2m by nla_id, date, and visit_no
* join to nla_id to create table nla
* no profile data for 12 nla07 and 160 nla12 lakes
* 184 lakes sampled twice
* 4 lakes sampled on visit_no == 2 only

```{r nla_data, include=FALSE, echo=FALSE, eval=FALSE} 
# get the profile data
  p07 <- read.csv(url("https://www.epa.gov/sites/production/files/2013-09/nla2007_profile_20091008.csv")) %>% 
    dplyr::select(nla_id = SITE_ID, visit_no = VISIT_NO, date = DATE_PROFILE, depth = DEPTH, 
                  temperature = TEMP_FIELD) %>% 
    dplyr::filter( !is.na(temperature)) %>%
    dplyr::mutate(date = format(mdy(date), format="%Y%m%d")) # 12648     5

   p12 <- read.csv(url("https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_profile_08232016.csv")) %>%
    filter(SAMPLE_TYPE == "PROF", SITE_ID != '', !is.na(DEPTH), !is.na(TEMPERATURE)) %>%
    dplyr::select(nla_id = SITE_ID, visit_no = VISIT_NO, date = DATE_COL, depth = DEPTH, 
                  temperature = TEMPERATURE) %>%
    dplyr::mutate(date = format(dmy(date), format="%Y%m%d")) # 10967     5

#rbind p07 & p12  
  prof <- rbind(p07, p12)
  p2m <- dplyr::filter(prof, depth <= 2) # 9045    5

  tmean_2m <- group_by(p2m, nla_id, visit_no, date) %>% 
    summarise(tmean_2m = mean(temperature), tmean_2m_n = n()) %>%
    left_join(nla_id) %>%
    dplyr::filter(!is.na(comid)) %>%
    dplyr::select(nla_id, comid, year, date, visit_no, tmean_2m, tmean_2m_n) %>%
    ungroup()

# some lakes have missing profile data  
  dim(tmean_2m) # 2370    7
  length(unique(tmean_2m$nla_id)) # 2186
  a <- filter(nla_id, !nla_id %in% tmean_2m$nla_id)
  table(a$year) # 2007 = 12; 2012 = 160

# some lakes sampled twice
  table(tmean_2m$visit_no) # 1 = 2182; 2 = 188

  a <- as.data.frame(table(tmean_2m$nla_id))
  b <- filter(a, Freq > 1) # 184 2; so 4 lakes sampled on visit_no == 2 only

# save tmean_2m
        saveRDS(object = tmean_2m, file = here::here('data/tmean_2m.Rds'))
```

### nlcd impervious data
* calculate impervious surface in a 3000m buffer  for lmorpho / NLA lakes
    - use year = 2006 for the 2007 lakes
    - use year = 2011 for the 2012 lakes
* restrict to lakes in tmean_2m since these have water temperature data.
* 3k buffer also used in the ecosphere paper: https://esajournals.onlinelibrary.wiley.com/doi/epdf/10.1002/ecs2.1321
* The buffering failed for comid == 14489102 (nla_id == NLA06608-0243 & nla_id == NLA12_SD-101); for now this lake will be removed.
* goatscape did not return data for 14 of the 2007 lakes and 2 of the 2012 lakes.  We could work on this further but it is easiest just to eliminate these.  
* The percent overlap field shows what percent of the buffer is covered by NLCD.  For 2007 and 2012 1 and 2 lakes are not completely covered respectively.  Eliminate these. 
* For QAQC percent impervious was calculated in ArcMap for a couple of lakes.  Results comparable.

```{r imperv, include=FALSE, echo=FALSE, eval=FALSE} 
if(!exists("tmean_2m")) tmean_2m <- readRDS(here::here('data/tmean_2m.Rds'))
if(!exists("lmorpho")) load(here::here('data_local/lmorpho.rda'))

#  albers projection used by nlcd
crs_alb <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 
  +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
   landscape<-sf::st_transform(landscape, crs_alb) #reproject to the Albers Equal Area Conic projection used by NLCD

# get the comids for nla lakes from tmean_2m
# remove comid == 14489102; it fails in the buffer step
  ids <- dplyr::select(tmean_2m, nla_id, comid, year) %>%   # 2370    3
          filter(comid != 14489102) %>%                     # 2367    3
          distinct()                                        # 2184    3

  filter(ids, year == 2007) %>% dplyr::select(comid) %>% distinct() %>% dim() # 1119    1
  filter(ids, year == 2012) %>% dplyr::select(comid) %>% distinct() %>% dim() # 1065    1; 1119 + 1065 = 2184

# get the % imperv
# set up data.frame for results
impervious <- data.frame(comid = as.numeric(rep(NA, length(ids$comid))),
                    nla_id = as.character(rep(NA, length(ids$comid))),
                    total_percent_impervious = as.numeric(rep(NA, length(ids$comid))),
                    total_areaM2_impervious = as.numeric(rep(NA, length(ids$comid))),
                    buf_area_m2 = as.numeric(rep(NA, length(ids$comid))),
                    percent_overlap = as.numeric(rep(NA, length(ids$comid))))

# loop to populate data.frame
for(i in min(which(is.na(impervious))):length(ids$comid)) {
  #for(i in min(which(is.na(impervious))):5) {
start <- Sys.time()
lago <- tryCatch(st_transform(filter(lmorpho, comid == ids$comid[i]), crs_alb), error = function(e) NA) 
buf <- tryCatch(sf::st_difference(sf::st_buffer(lago, 3000), sf::st_geometry(lago)), error = function(e) NA) 
imp <- tryCatch(gs_nlcd(buf, spatial = TRUE, year = ids$year[i]-1, dataset = "impervious"), 
                error = function(e) NA)

# harvest data
impervious$comid[i] <- ids$comid[i]
impervious$nla_id[i] <- ids$nla_id[i]

if(!is.na(imp)[1]){
  impervious$total_percent_impervious[i] <- imp$total_percent_impervious
  impervious$total_areaM2_impervious[i] <- imp$total_areaM2_impervious
  impervious$buf_area_m2[i] <- sum(imp$nlcd_freq$areaM2)
  impervious$percent_overlap[i] <- imp$percent_overlap
}

plot(buf[1], main = paste0("i = ", i, " of ", nrow(ids), "; nla_id = ", ids$nla_id[i], "; end = ", 
                           format(Sys.time(), "%H:%M:%OS"))) 
print(i)
print(Sys.time()-start)
}

# remove observations where is.na(total_percent_impervious) or percent_overlap < 100
dim(impervious) # 2184    6
impervious <- filter(impervious, !is.na(impervious$total_percent_impervious)) # 2168    6
impervious <- filter(impervious, percent_overlap == 100) #  2168    6

# save impervious
    saveRDS(object = impervious, file = here::here('data/impervious.Rds'))
    # impervious <- readRDS(file = here::here('data/impervious.Rds'))

```

* create table "nla"
    * primary keys are nla_id & visit_no

```{r tbl_nla, include=FALSE, echo=FALSE, eval=FALSE} 
tmean_2m <- readRDS(file = here::here('data/tmean_2m.Rds')) # 2370 7
# tmean_2m$date <- as.numeric(tmean_2m$date)

impervious <- readRDS(file = here::here('data/impervious.Rds')) # 2168    6
imp <- dplyr::select(impervious, nla_id, comid, 
                                         percent_impervious = total_percent_impervious, 
                                         area_impervious = total_areaM2_impervious, area_total = buf_area_m2)

nla <- left_join(tmean_2m, imp) # 2370    10
  dim(unique(dplyr::select(nla, nla_id, visit_no))) # 2370 2

# connect to the database
  means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

# add nla to table nla
  dbWriteTable(means, "nla", nla, overwrite = TRUE)

# index table nla
  dbExecute(means,"CREATE INDEX index_nla_comid ON nla (comid)") 
      #  dbExecute(means,"DROP INDEX index_nla_comid")
  dbExecute(means,"CREATE INDEX index_nlaid_comid ON nla (nla_id)")

# close and save the database
  dbDisconnect(means)
```

### QAQC tmeans_raw: checked and ready to go
* Use arcMap to extract the prism values for 1 date and compare to table tmean_raw
    * open arcMap and add prism raster tmeans for 19920930 ('data_local/prism_qaqc.mxd')
    * save table "lakes" as lakes.csv
    * import into arcmap and save as lakes.shp
    * use the extract multi point tools to add prism tmean to lakes.shp (.dbf)
    * bring back into r and compare
    * everything peachy: 
      * median difference = 0.0000006
      * max difference = 2.7 # checked this out and the lake is on the border of two grid cells
* no data for 16 cellnums corresponding to 22 lakes 
    * looked at these in arcmap and all are just outside the prism raster; no data

```{r tm_raw, include=FALSE, echo=FALSE, eval=FALSE} 
# create a shapefile from table "lakes"
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
dbListFields(means, "tmeans_raw")

lakes <- dbGetQuery(means, 'SELECT lakes.* FROM "lakes"') # 363300     14
  write.csv(lakes, file = here::here('data_local/lakes.csv'), row.names = FALSE)

# import into arcmap and save as shapefile 'data_local/lakes.shp'
# compare to tmeans_raw
am <- read.csv(here::here('data_local/lakes.dbf.csv')) # 363300      5
db <- dbGetQuery(means, 'SELECT cellnum, d0930 FROM "tmeans_raw" WHERE year = 1992') # 144402      2
comp <- left_join(am, db) # 363300      6
comp <- comp[complete.cases(comp), ] # 363278      6
comp$dif <- abs(comp$d0930 - comp$tm19920930)
comp <- arrange(comp, dif)
summary(comp$dif)

plot(comp$d0930, comp$tm19920930)

# missing values
table(is.na(db$cellnum)) # FALSE = 144402
table(is.na(db$d0930)) # FALSE = 144386 TRUE = 16

chk <- db$cellnum[which(is.na(db$d0930))]

# close and save the database
dbDisconnect(means)
```

### QAQC tmeans
* A date chosen randomly
* Get the raw prism data from table tmean_raw for the date and the 30 days prior for all cellnums
* Re-calculate the values in table tmeans from the raw data for all cellnums
* Compare the results to table tmeans
    * Done several times - all values match
* Re-extract the values for the date from the prism raster and compare
    * Also done several times - all values match
    * This also verifies the cellnum field in table 'lakes'

```{r tm, include=FALSE, echo=FALSE, eval=FALSE} 
# choose a random date
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
dbListFields(means, "tmeans")

# list of all dates
dates <- dbGetQuery(means, 'SELECT DISTINCT date FROM "tmeans"') # 4514    1

# choose a date at random 
d <- sample(dates$date, 1)
yr <- substr(d, 1, 4)
fields <- c('cellnum', 'year', paste0("d", substr(format(seq(ymd(d)-30, ymd(d), by="days"), format="%Y%m%d"), 5, 8)))

# get tmean_raw data for date
tmr <- dbGetQuery(means, paste("SELECT * FROM 'tmeans_raw' WHERE year = ", yr))
tmr <- tmr[ , fields] # 144402     33

comp <- dplyr::select(tmr, cellnum)
comp$date <- d
comp$tmean <- tmr[ , fields[33]]
comp$tmean_dm1 <- tmr[ , fields[32]]
comp$tmean_avg3 <- apply(tmr[, 30:32], 1, mean)
comp$tmean_avg7 <- apply(tmr[, 26:32], 1, mean)
comp$tmean_avg30 <- apply(tmr[, 3:32], 1, mean)

aves <- dbGetQuery(means, paste0("SELECT * FROM 'tmeans' WHERE date = ", d)) # 144402      7

all.equal(comp$cellnum, aves$cellnum) # TRUE
all.equal(comp$tmean, aves$tmean) # TRUE
all.equal(comp$tmean_dm1, aves$tmean_dm1) # TRUE
all.equal(comp$tmean_avg3, aves$tmean_avg3) # TRUE
all.equal(comp$tmean_avg7, aves$tmean_avg7) # TRUE
all.equal(comp$tmean_avg30, aves$tmean_avg30) # TRUE

# check that the cellnum field in table "lakes" is correct
# get the tmeans_raw data for the date
tmrd <- dplyr::select(tmr, cellnum, paste0('d', substr(d, 5, 8)))

# get the prism tmeans raster values for the date
pr <- raster(grep(paste0("PRISM_tmean_stable_4kmD1_", d), ls_prism_data(absPath=T)[,2],value=T))
prv <- data.frame(value = values(pr))
prv$cellnum <- as.numeric(row.names(prv))

# join and compare
chk <- left_join(tmrd, prv)
all.equal(chk[ ,2], chk[ ,3]) # TRUE

# close and save the database
dbDisconnect(means)
```

### QAQc impervious
* repeat the impervious buffer clip in arcmap
* join the nla to lmorpho and transform to nlcd crs
* write nla lakes to 'data_local/nla.shp'
* download 2011 % Dev. Imperv. CONUS: (https://www.mrlc.gov/data?f%5B0%5D=category%3Aurban%20imperviousness)
* add nla.shp and NLCD imperv to arcmap ('data_local/nlcd_qaqc.mxd)
* select nla 2012 lakes for visit_no == 2 
* buffer with "buffer" tool 
    * Distance = linear unit = 3000m
    * Side type = Outside Only
    * End Type = Round
    * Method = Planar
    * Dissolve Type = None
    * save as: nla2012_visit2_buf3000
* use the clip (Data Management) tool to clip 2011 nlcd raster to the buffer 
    * input raster = nlcd imp 2011
    * output extent = nla2012_visit2_buf3000
    * Check Box: Use input Features for Clipping Geometry = Checked
    * NoData = 255
    * save as: nla12_v1_buf3000_imp.tif
* use Zonal Statistics as Table to sumarize raster clip
    * input ... zone data = nla2012_visit2_buf3000
    * zone field = 'nla_id'
    * nla12_v1_buf3000_imp.tif
    * output table = nla12_v1_buf3000_imp.dbf
* compare arcmap and nla.impervious values graphically - near perfect fit.  Small differences expected.

```{r chk_imp, include=FALSE, echo=FALSE, eval=FALSE} 
# get the nlcd crs
a <- raster::raster(here::here('nlcd_data/cropped/buf_NLCD_2006_impervious.tif'))
crs_nlcd <- as.character(raster::crs(a))

# join nla and lmorpho
load(here::here('data_local/lmorpho.rda'))
  names(lmorpho)[1] <- 'comid'
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
  nla <- dbGetQuery(means, 'SELECT nla.* FROM "nla"') # 2370    9
        dbDisconnect(means)

names_nla <- names(nla)

nla <- left_join(lmorpho, nla) %>%
    dplyr::filter(!is.na(nla_id)) %>%
    dplyr::select(names_nla) %>% # 2370   10
    sf::st_transform(crs_nlcd)

# st_write(nla, here::here('data_local/nla.shp'), delete_layer = TRUE)

# join nla & arcmap summary nla12_v1_buf3000_imp.dbf
am12 <- foreign::read.dbf(here::here("data_local/nla12_v1_buf3000_imp.dbf"))
chk_am12 <- left_join(am12, nla) %>% 
          dplyr::select(nla_id, arc_per_imp = MEAN, percent_impervious, arc_area = AREA, area_total)

# compare by plotting
plot(chk_am12$arc_per_imp, chk_am12$percent_impervious)
  abline(0, 1)

plot(chk_am12$arc_area, chk_am12$area_total)
  abline(0, 1)
```

### QAQC leftovers
* The database elevation compared to the nla design file elevations; excellent match
* tmean_n recalculated in excel and compared to database; all.equal == TRUE

```{r chk_nla, include=FALSE, echo=FALSE, eval=FALSE} 
######### elevation
# get the the [elevatr] elevations
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
  nla <- dbGetQuery(means, 'SELECT nla.nla_id, lakes.elevation, tmean_2m, tmean_2m_n 
                          FROM "nla" 
                          INNER JOIN "lakes"
                          ON nla.comid = lakes.comid
                          WHERE nla.visit_no = 1') # 2182    4
              dbDisconnect(means)

# get the lake elevations from the design files
n07 <- read.csv(
    url("https://www.epa.gov/sites/production/files/2014-01/nla2007_sampledlakeinformation_20091113.csv"))

n07a <- filter(n07, VISIT_NO == 1) %>% 
          dplyr::select(nla_id = SITE_ID, nla_elev = ELEV_PT)

n12 <- read.csv(
    url("https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_siteinfo_08232016.csv"))

n12a <- filter(n12, VISIT_NO == 1) %>% 
          dplyr::select(nla_id = SITE_ID, nla_elev = ELEVATION)

nla_elev <- rbind(n07a, n12a)

# join elevation files
elev <- left_join(nla, nla_elev)

# plot to compare
plot(elev$elevation, elev$nla_elev)
  abline(0, 1, col = "red", lwd =3)

######## tmean_2m

# get the profile data
  p07 <- read.csv(url("https://www.epa.gov/sites/production/files/2013-09/nla2007_profile_20091008.csv")) %>% 
    dplyr::select(nla_id = SITE_ID, visit_no = VISIT_NO, date = DATE_PROFILE, depth = DEPTH, 
                  temperature = TEMP_FIELD) %>% 
    dplyr::filter( !is.na(temperature)) %>%
    dplyr::mutate(date = format(mdy(date), format="%Y%m%d")) # 12648     5

   p12 <- read.csv(url("https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_profile_08232016.csv")) %>%
    filter(SAMPLE_TYPE == "PROF", SITE_ID != '', !is.na(DEPTH), !is.na(TEMPERATURE)) %>%
    dplyr::select(nla_id = SITE_ID, visit_no = VISIT_NO, date = DATE_COL, depth = DEPTH, 
                  temperature = TEMPERATURE) %>%
    dplyr::mutate(date = format(dmy(date), format="%Y%m%d")) # 10967     5

# rbind p07 & p12  
  prof <- rbind(p07, p12)
  p2m <- dplyr::filter(prof, depth <= 2, visit_no == 1) # 8322    5

# write to csv
  write.csv(p2m, here::here("data_local/ptm.csv"), row.names = FALSE)

# used excel pivot table to recalculate the means.  
# compare to nla

excel <- read.csv(here::here("data_local/ptm.csv")) %>%
  dplyr::select(nla_id = nla_id_sum, temp_mean, temp_n)

tcomp <- inner_join(nla, excel)
all.equal(tcomp$tmean_2m, tcomp$temp_mean)  # TRUE
all.equal(tcomp$tmean_2m_n, tcomp$temp_n)  # TRUE
```

## Document database
* create a table in the database of data definitions (data_def) from 'data/lake_temperatures.xlsx'

```{r doc_db, include=FALSE, echo=FALSE, eval=TRUE, cache=TRUE} 
dd <- read_xlsx(here::here('data/lake_temperatures_sqlite.xlsx'), sheet = "data_defs") %>%
  dplyr::select(-rmd)

# add table data_def to the database
means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))
dbWriteTable(means, "data_defs", dd, overwrite = TRUE)

tables <- dbListTables(means)
tables

# data_defs
  dd_f <- dbListFields(means, "data_defs")  # fields
  dd_n <- dbGetQuery(means, 'SELECT Count(*) FROM data_defs') # 46 observations

# lakes
  l_f <- dbListFields(means, "lakes")  # fields
  l_n <- dbGetQuery(means, 'SELECT Count(*) FROM lakes') # 363300 observations
  dim(dbGetQuery(means, 'SELECT DISTINCT comid FROM lakes')) # 363300; primary key

# nla
  n_f <- dbListFields(means, "nla")  # fields
  n_n <- dbGetQuery(means, 'SELECT Count(*) FROM nla') # 2370 observations
  dim(dbGetQuery(means, 'SELECT DISTINCT nla_id, visit_no FROM nla')) # 2370; composite primary key

# tmeans
  tm_f <- dbListFields(means, "tmeans")  # fields
  tm_n <- dbGetQuery(means, 'SELECT Count(*) FROM tmeans') # 651830628 (651.8 m) observations
  dim(dbGetQuery(means, 'SELECT DISTINCT cellnum FROM tmeans')) # 144402; composite primary key
  dim(dbGetQuery(means, 'SELECT DISTINCT date FROM tmeans')) # 4514; composite primary key
    # NOTE:  144402 * 4514 = 651830628

# tmeans_raw
  tmr_f <- dbListFields(means, "tmeans_raw")  # fields
  tmr_n <- dbGetQuery(means, 'SELECT Count(*) FROM tmeans_raw') # 5342874 (5.3 m) observations
  dim(dbGetQuery(means, 'SELECT DISTINCT cellnum FROM tmeans_raw')) # 144402; composite primary key
  dim(dbGetQuery(means, 'SELECT DISTINCT year FROM tmeans_raw')) # 37; composite primary key
    # NOTE:  144402 * 37 = 5342874

# close and save the database
dbDisconnect(means)

# database size
size <- paste(round(file.size(here::here('data_local/lake_temperatures.sqlite')) / 1073741824, 1), 'gb')
```

* The database ('data_local/lake_temperatures.sqlite') is too large (`r size`) to keep on github.  It will have to be compiled by running the code chunks in this document.  It could be a lengthy and difficult task.

* The database includes the `r length(tables)` tables: `r tables`

#### Data definitions for table "data_defs"  
* This table has the data definitions for all tables in [lake_temperatures.sqlite]
* observations = `r dd_n` 
* fields = `r dd_f`

field	| 	description
----------- | --------------------------------- 
**table**|database table for data definitions
**field**|field in database table for data definitions
**source**|source of the data for the field
**units**|units of measurement
**indexed**|does a database index exist for the field
**primary**|Is this field a primary field?  NA = no; primary = simple primary key; composite = primary key composed of 2 or more fields
**description**|a general description of the field

#### Data definitions for table "lakes" 
* This table has the morphometry and elevation for the lmorpho lakes
* observations = `r l_n` 
* fields = `r l_f`

field	| source | units | indexed |	primary |	description
----------- | -----------	| ------- | ---- | ---- |	--------------------------------- 
**cellnum**|prism|NA|YES||the grid cell number of the prism raster(s) for the lake
**comid**|lmorpho|int|YES|primary|nhdplus comid from the lmorpho dataset
**longitude**|lmorpho|dd|||longitude of lmorpho centroid. Transformed to prism crs =="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
**latitude**|lmorpho|dd|||latitude of lmorpho centroid. Transformed to prism crs =="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
**surface_area**|lmorpho|m2|||lake surface area
**shoreline_length**|lmorpho|m |||length of lake shoreline
**shoreline_dev**|lmorpho|NA|||shoreline development index
**max_length**|lmorpho|m|||maximum length of the lake polygon
**max_width**|lmorpho|m|||maximum width of the lake polygon
**mean_width**|lmorpho|m|||mean width of the lake polygon
**max_depth**|lmorpho|m|||estimated maximum depth of lake
**mean_depth**|lmorpho|m|||estimated mean depth of lake
**volume**|lmorpho|m3|||estimated volume depth of lake
**elevation**|aws|m|||elevation of the lake

#### Data definitions for table "nla" 
* This table has the NLA water temperature and NLCD impervious data
* observations = `r n_n` 
* fields = `r n_f`

field	| source | units | indexed |	primary |	description
----------- | -----------	| ------- | ---- | ---- |	--------------------------------- 
**nla_id**|nla|char|YES|composite|unique identifiers for the NLA lakes
**comid**|lmorpho|int|YES||nhdplus comid from the lmorpho dataset
**cellnum**|prism|NA|||the grid cell number of the prism raster(s) for the lake
**year**|nla|YYYY|||sample year 2007 or 2012
**date**|nla|YYYYMMDD|||date the profile temperatures were collected
**visit_no**|nla|int||composite|All lakes were sampled once (visit_no = 1); some were sampled twice (visit_no = 2)
**tmean_2m**|nla|degrees C|||mean water temperature for depth <= 2 m
**tmean_2m_n**|nla|degrees C|||number of observations used to calculate tmean_2m
**percent_impervious**|nlcd|percent|||percent impervious surface in a 3000m buffer around the lake (includes islands); only observations with 100% coverage kept (i.e., lakes with partial NLCD coverage such as border lakes are excluded)
**area_impervious**|nlcd|m2|||m2 of impervious surface in a 3000m buffer around the lake (includes islands); only observations with 100% coverage kept (i.e., lakes with partial NLCD coverage such as border lakes are excluded)
**area_total**|lmorpho|m2|||area of 3000 m buffer around the lmorpho lake

#### Data definitions for table "tmeans" 
* This table has the aggregated prism air temperature estimates
* observations = `r tm_n` 
* fields = `r tm_f`

field	| source | units | indexed |	primary |	description
----------- | -----------	| ------- | ---- | ---- |	--------------------------------- 
**cellnum**|prism|NA|YES|composite|the grid cell number of the prism raster(s) for the lake
**date**|prism|YYYYMMDD|YES|composite|date of the prism estimate
**tmean**|prism|degrees C|||mean temperature on the date
**tmean_dm1**|prism|degrees C|||mean temperature on the day prior to the date (date minus 1)
**tmean_avg3**|prism|degrees C|||average of prism means for the 3 days prior to the date
**tmean_avg7**|prism|degrees C|||average of prism means for the 7 days prior to the date
**tmean_avg30**|prism|degrees C|||average of prism means for the 30 days prior to the date

#### Data definitions for table "tmeans_raw" 
* This table has the raw (by date) prism air temperature estimates
* observations = `r tmr_n` 
* fields = `r tmr_f`

field	| source | units | indexed |	primary |	description
----------- | -----------	| ------- | ---- | ---- |	--------------------------------- 
**cellnum**|prism|NA|YES|composite|the grid cell number of the prism raster(s) for the lake
**d0501:d0531**|prism|degrees C|||estimated mean temperature for dates May 1:May 31 by tmeans_raw.year
**d0601:d0630**|prism|degrees C|||estimated mean temperature for dates June 1:June 30 by tmeans_raw.year
**d0701:d0731**|prism|degrees C|||estimated mean temperature for dates July 1:July 31 by tmeans_raw.year
**d0801:d0831**|prism|degrees C|||estimated mean temperature for dates August 1:August 31 by tmeans_raw.year
**d0901:d0930**|prism|degrees C|||estimated mean temperature for dates September 1:September 30 by tmeans_raw.year
**year**|prism|YYYY|YES|composite|year 1981 to 2017

## Create a subset of the data for lake temperature model and save as nla_base.Rds

# Set up Data
* load nla data from "data_local/lake_temperatures.sqlite"
* create nla_base with the 2007 and 2012 nla, prism, lmorpho & imperv data for random forest analysis

```{r bdata, include=FALSE, echo=FALSE, cache=TRUE}
# load the nla data from lake_temperatures.sqlite

if(file.exists(here::here('data/nla_base.Rds'))) {
  nla_base <- readRDS(here::here('data/nla_base.Rds')) 
  } else {

means <- dbConnect(RSQLite::SQLite(), here::here("data_local/lake_temperatures.sqlite"))

# get the nla_base data
nla_base <- dbGetQuery(means, '
  SELECT nla.nla_id, nla.comid, lakes.cellnum, nla.year, nla.date, nla.visit_no, 
    nla.tmean_2m, nla.percent_impervious, lakes.longitude, lakes.latitude, lakes.surface_area, 
    lakes.shoreline_length, lakes.shoreline_dev, lakes.max_length, lakes.max_depth, lakes.max_width, 
    lakes.mean_width, lakes.elevation, lakes.mean_depth, lakes.volume
  FROM nla INNER JOIN lakes ON nla.comid = lakes.comid')

# add fields for tmeans
nla_base$tmean <- NA
nla_base$tmean_dm1 <- NA
nla_base$tmean_avg3 <- NA
nla_base$tmean_avg7 <- NA
nla_base$tmean_avg30 <- NA

for(i in min(which(is.na(nla_base$tmean))):nrow(nla_base)){
  print(i)
  print(Sys.time())
tm <- dbGetQuery(means, paste0("SELECT * FROM 'tmeans' WHERE cellnum = ", nla_base$cellnum[i], 
                         " AND date = ", nla_base$date[i]))

if(nrow(tm) > 0) {
  nla_base$tmean[i] <- tm$tmean
  nla_base$tmean_dm1[i] <- tm$tmean_dm1
  nla_base$tmean_avg3[i] <- tm$tmean_avg3
  nla_base$tmean_avg7[i] <- tm$tmean_avg7
  nla_base$tmean_avg30[i] <- tm$tmean_avg30
}}

dbDisconnect(means)

saveRDS(nla_base, file = here::here('data/nla_base.Rds')) # 2370   25
write.csv(nla_base, here::here('data/nla_base.csv'), row.names = FALSE)
}
```

### data definitions for nla_base.Rds & nla_base.csv

field	| source | units | indexed |	primary |	description
----------- | -----------	| ------- | ---- | ---- |	--------------------------------- 
**nla_id**|nla|char|YES|composite|unique identifiers for the NLA lakes
**comid**|lmorpho|int|YES||nhdplus comid from the lmorpho dataset
**cellnum**|prism|NA|||the grid cell number of the prism raster(s) for the lake
**year**|nla|YYYY|||sample year 2007 or 2012
**date**|nla|YYYYMMDD|||date the profile temperatures were collected
**visit_no**|nla|int||composite|All lakes were sampled once (visit_no = 1); some were sampled twice (visit_no = 2)
**tmean_2m**|nla|degrees C|||mean water temperature for depth <= 2 m
**percent_impervious**|nlcd|percent|||percent impervious surface in a 3000m buffer around the lake (includes islands); only observations with 100% coverage kept (i.e., lakes with partial NLCD coverage such as border lakes are excluded)
**longitude**|lmorpho|dd|||longitude of lmorpho centroid. Transformed to prism crs =="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
**latitude**|lmorpho|dd|||latitude of lmorpho centroid. Transformed to prism crs =="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
**surface_area**|lmorpho|m2|||lake surface area
**shoreline_length**|lmorpho|m |||length of lake shoreline
**shoreline_dev**|lmorpho|NA|||shoreline development index
**max_length**|lmorpho|m|||maximum length of the lake polygon
**max_depth**|lmorpho|m|||estimated maximum depth of lake
**max_width**|lmorpho|m|||maximum width of the lake polygon
**mean_width**|lmorpho|m|||mean width of the lake polygon
**elevation**|aws|m|||elevation of the lake
**mean_depth**|lmorpho|m|||estimated mean depth of lake
**volume**|lmorpho|m3|||estimated volume depth of lake
**tmean**|prism|degrees C|||mean temperature on the date
**tmean_dm1**|prism|degrees C|||mean temperature on the day prior to the date (date minus 1)
**tmean_avg3**|prism|degrees C|||average of prism means for the 3 days prior to the date
**tmean_avg7**|prism|degrees C|||average of prism means for the 7 days prior to the date
**tmean_avg30**|prism|degrees C|||average of prism means for the 30 days prior to the date

### 2019-04-26 R/varsel_plot.R

@@ -1,0 +1,23 @@
+# https://raw.githubusercontent.com/USEPA/LakeTrophicModelling/master/R/varsel_plot.R
+
+#' Plot varsel_regression output
+#' 
+#' @param vs_reg_obj Object from output of \link{varsel_regression_rf}.
+#' @export
+#' @import ggplot2
+varsel_plot <- function(vs_reg_obj){
+  dat <- data.frame(num = vs_reg_obj$num_var, mse = vs_reg_obj$mse)
+  plot_out<-ggplot(dat, aes(x=num,y=mse))+geom_point(size=2.5)+
+    theme(text = element_text(family="sans"),
+          panel.background = element_blank(), #panel.grid = element_blank(), 
+          panel.border = element_rect(fill = NA), 
+          plot.title  = element_text(family="sans",size=15,face="bold",vjust=1.1),
+          legend.key = element_rect(fill = 'white'),
+          legend.text = element_text(family="sans",size=15), legend.title = element_text(size=15),
+          axis.title.x = element_text(family="sans",vjust = -0.5, size = 12),
+          axis.title.y = element_text(family="sans",vjust = 1.5, size = 12),
+          axis.text.x = element_text(family="sans",size = 11),
+          axis.text.y = element_text(family="sans",size = 11))+
+    labs(x="Number of variables in model", y="Root Mean Square Error")
+  return(plot_out)
+}

### 2019-04-26 R/varsel_regression_rf.R

@@ -1,0 +1,46 @@
+# https://raw.githubusercontent.com/USEPA/LakeTrophicModelling/master/R/varsel_regression_rf.R
+
+#' Variable Selection with Regression RF
+#' 
+#' Similar approach to varSelRF, but for regresssion.  Use full model to rank
+#' variables based on %IncMSE or gini. Step through that list with most important first
+#' run RF, store variables and %MSE of model.
+#' @param y response a vector 
+#' @param x predictors a data.frame
+#' @param ... options to pass to randomForest; default (importance = FALSE) ranks on gini; to rank on %IncMSE
+#' use "importance = TRUE".
+#' @export
+#' @import randomForest
+#' @examples
+#' data(LakeTrophicModelling)
+#' predictors_all <- predictors_all[predictors_all!="DATE_COL"]
+#' all_dat <- data.frame(ltmData[predictors_all],LogCHLA=log10(ltmData$CHLA))
+#' all_dat <- all_dat[complete.cases(all_dat),]
+#' x<-varsel_regression_rf(all_dat$LogCHLA,all_dat[,names(all_dat)!="LogCHLA"],
+#'                         ntree=100,prog=T)
+varsel_regression_rf <- function(y,x,prog=F,...){
+  out <- list(mse=NULL,rsq=NULL,num_var=NULL,vars=NULL)
+  dat <- data.frame(y=y,x)
+  init_rf <- randomForest(y=dat$y,x=dat[,names(x)],...)
+  init_imp <- importance(init_rf)
+  var_sort <- rownames(init_imp)[order(init_imp[,1],decreasing = TRUE)]
+  vars <- NULL
+  for(i in var_sort){
+    if(is.null(vars)){ 
+      vars <- c(vars,i) 
+      idx <- 1
+    } else {
+      vars <- c(vars,i)
+      vars_rf <- randomForest(y=dat$y,x=dat[,vars],...)
+      out$mse[[idx]] <- vars_rf$mse[length(vars_rf$mse)]
+      out$rsq[[idx]] <- vars_rf$rsq[length(vars_rf$rsq)]
+      out$num_var[[idx]] <- length(vars)
+      out$vars[[idx]] <- vars
+      idx <- idx + 1
+    }
+    if(prog & idx%%3==0){
+      print(paste0(round(idx/length(var_sort)*100,1),"% completed"))
+    }
+  }
+  return(out)
+}

### 2019-04-30 R/model.Rmd
refining the 2007&2012 model
Change variable importance to MSE

@@
 library(ggplot2)
+library(lubridate)
 ```
@@
 
+##Reading in and creating data sets
+nla<-read.csv(here::here('data/nla_base.csv'))
 x<-complete.cases(nla)
@@
 nla.cc$year<-factor(nla.cc$year)
+nla.cc$date <- lubridate::yday(lubridate::ymd(nla.cc$date))
 
@@
             shoreline_length, shoreline_dev, max_length, max_width, mean_width,
-            max_depth, max_depth, volume, elevation) # 1214   19
+            max_depth, max_depth, volume, elevation)
 
@@
 
-nla.all <-  select(nla.cc, tmean_2m, year, date, longitude, latitude, tmean,
-                   tmean_dm1, tmean_avg3, tmean_avg7, tmean_avg30,percent_impervious,                    surface_area,shoreline_length, shoreline_dev, max_length,                            max_width, mean_width, max_depth, max_depth, volume, elevation)
+nla.all <-  filter(nla.cc, tmean_2m < 70) %>%
+  dplyr::select(tmean_2m, year,date, longitude, latitude, tmean,          tmean_avg30,percent_impervious,surface_area,shoreline_length, shoreline_dev, max_length, max_width, mean_width, max_depth, max_depth, volume, elevation)
 
-var_sel_tmean07 <- varsel_regression_rf(data07$tmean_2m, select(data07, -tmean_2m),                ntree=1000, prog=TRUE)
+##variable selection for 2007-only model
+var_sel_tmean07 <- varsel_regression_rf(data07$tmean_2m, select(data07, -tmean_2m), ntree=1000, prog=TRUE, importance=TRUE)
 
-varsel_plot(var_sel_tmean07) #9 variables
+varsel_plot(var_sel_tmean07)
 
@@
 
-var_sel_tmean12 <- varsel_regression_rf(data12$tmean_2m, select(data12, -tmean_2m),                ntree=1000, prog=TRUE)
+##Variable selection for 2012-only model
+var_sel_tmean12 <- varsel_regression_rf(data12$tmean_2m, select(data12, -tmean_2m),                ntree=1000, prog=TRUE,importance=TRUE)
 
@@
 
-var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE)
+##Variable selection for 2007&2012 model
+var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
 
-varsel_plot(var_sel_tmean) #9 variables
+varsel_plot(var_sel_tmean) #7 variables
 
-var_sel_tmean$vars[[9]]
+var_sel_tmean$vars[[7]]
 
@@
 
-  select(tmean_2m,var_sel_tmean$vars[[9]]) %>%
+  select(tmean_2m,var_sel_tmean$vars[[7]]) %>%
 

### 2019-05-03 R/model.Rmd
added set.seed

@@
 
-data07 <- filter(nla.cc, year == 2007, tmean_2m < 70) %>%
-            dplyr::select(tmean_2m, date, longitude, latitude, tmean, tmean_dm1,
-            tmean_avg3, tmean_avg7, tmean_avg30, percent_impervious, surface_area,
-            shoreline_length, shoreline_dev, max_length, max_width, mean_width,
-            max_depth, max_depth, volume, elevation)
-
-data12 <- filter(nla.cc, year == 2012, tmean_2m < 70) %>%
-            dplyr::select(tmean_2m, date, longitude, latitude, tmean, tmean_dm1,
-            tmean_avg3, tmean_avg7, tmean_avg30, percent_impervious, surface_area,
-            shoreline_length, shoreline_dev, max_length, max_width, mean_width,
-            max_depth, max_depth, volume, elevation)
+# data07 <- filter(nla.cc, year == 2007, tmean_2m < 70) %>%
+#             dplyr::select(tmean_2m, date, longitude, latitude, tmean, tmean_dm1,
+#             tmean_avg3, tmean_avg7, tmean_avg30, percent_impervious, surface_area,
+#             shoreline_length, shoreline_dev, max_length, max_width, mean_width,
+#             max_depth, max_depth, volume, elevation)
+#             
+#             
+# data12 <- filter(nla.cc, year == 2012, tmean_2m < 70) %>%
+#             dplyr::select(tmean_2m, date, longitude, latitude, tmean, tmean_dm1,
+#             tmean_avg3, tmean_avg7, tmean_avg30, percent_impervious, surface_area,
+#             shoreline_length, shoreline_dev, max_length, max_width, mean_width,
+#             max_depth, max_depth, volume, elevation)
 
 nla.all <-  filter(nla.cc, tmean_2m < 70) %>%
-  dplyr::select(tmean_2m, year,date, longitude, latitude, tmean,          tmean_avg30,percent_impervious,surface_area,shoreline_length, shoreline_dev, max_length, max_width, mean_width, max_depth, max_depth, volume, elevation)
+              dplyr::select(tmean_2m, year,date, longitude, latitude, tmean,
+              tmean_avg30,percent_impervious,surface_area,shoreline_length,
+              shoreline_dev, max_length, max_width, mean_width, max_depth,
+              max_depth, volume, elevation)
 
 ##variable selection for 2007-only model
-var_sel_tmean07 <- varsel_regression_rf(data07$tmean_2m, select(data07, -tmean_2m), ntree=1000, prog=TRUE, importance=TRUE)
-
-varsel_plot(var_sel_tmean07)
-
-var_sel_tmean07$vars[[9]]
-
-##Variable selection for 2012-only model
-var_sel_tmean12 <- varsel_regression_rf(data12$tmean_2m, select(data12, -tmean_2m),                ntree=1000, prog=TRUE,importance=TRUE)
-
-varsel_plot(var_sel_tmean12) #9 variables
-
-var_sel_tmean12$vars[[9]]
+# var_sel_tmean07 <- varsel_regression_rf(data07$tmean_2m, select(data07, -tmean_2m), ntree=1000, prog=TRUE, importance=TRUE)
+# 
+# varsel_plot(var_sel_tmean07) 
+# 
+# var_sel_tmean07$vars[[9]]
+# 
+# 
+# ##Variable selection for 2012-only model
+# var_sel_tmean12 <- varsel_regression_rf(data12$tmean_2m, select(data12, -tmean_2m),                ntree=1000, prog=TRUE,importance=TRUE)
+# 
+# varsel_plot(var_sel_tmean12) #9 variables
+# 
+# var_sel_tmean12$vars[[9]]
 
 ##Variable selection for 2007&2012 model
+set.seed(42)
 var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
@@
 
-RF2007 <- data07 %>%
+# RF2007 <- data07 %>%
+# 
+#   select(tmean_2m,var_sel_tmean07$vars[[9]]) %>%
+# 
+#   randomForest(tmean_2m ~ ., data = ., importance = TRUE, 
+# 
+#                ntree = 10000)
+# 
+# 
+# 
+# RF2007
+# 
+# varImpPlot(RF2007)
+# 
+# 
+# 
+# RF2012 <- data12 %>%
+# 
+#   select(tmean_2m,var_sel_tmean12$vars[[9]]) %>%
+# 
+#   randomForest(tmean_2m ~ ., data = ., importance = TRUE, 
+# 
+#                ntree = 10000)
+# 
+# 
+# 
+# RF2012
+# 
+# varImpPlot(RF2012)
+#
 
-  select(tmean_2m,var_sel_tmean07$vars[[9]]) %>%
-
-  randomForest(tmean_2m ~ ., data = ., importance = TRUE,
-
-               ntree = 10000)
-
-RF2007
-
-varImpPlot(RF2007)
-
-RF2012 <- data12 %>%
-
-  select(tmean_2m,var_sel_tmean12$vars[[9]]) %>%
-
-  randomForest(tmean_2m ~ ., data = ., importance = TRUE,
-
-               ntree = 10000)
-
-RF2012
-
-varImpPlot(RF2012)
-
+set.seed(42)
 RFAll <- nla.all %>%

### 2019-05-06 manuscript/manuscript.Rmd
Adding text to intro and methods

@@
 
-During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world (Williamson et al.
-2009, Schindler 2009).
+During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].
 As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)(NASA GISS), Surface temperatures of lakes are also increasing globally (0.34 per decade from 1985 to 2009) [@o2015rapid].
@@
 Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.
-Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes (Downing et al.
-2006, Winslow et al.
-2014).
+Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes [@downing2006global; @winslow2014lake].
+
+Therefore, we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
+Modelling lake temperature requires a large amount of data, and therefore study lakes are often selected opportunistically, which may introduce a spatial bias.
+Frequently, the study lakes have a high regional resource value (e.g.
+the Great Lakes) and commonly have an extensive monitoring history due  to the vested interest of the public.
+When model efforts do attempt to cross large spatial extents, these efforts often rely on satellite data.
+While these models predict over large areas, they are restricted by the size of the lake captured by satellite (typically 3km2 for 1 km MODIS pixels).
+
+The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
+The NLA is stratified random sample of all lakes in the conterminous United State.
+The NLA was conducted in 2007 and then repeated in 2012.
+Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.
+The uniqueness of this data set allows us to build a robust lake photic zone temperature model for all US lakes.
+
+The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
+The NLA is stratified random sample of all lakes in the conterminous United States.
+The NLA was conducted in 2007 and then repeated in 2012.
+Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.
+The uniqueness of this data set allows us to build a robust lake photic zone temperature model for all US lakes.
+By using lakes across the US (i.e.
+at a large spatial extent), we included lakes with different morphologies, in different climates with diverse geologies, and surrounding landscapes.
+Additionally, small lakes were well represented with more than 50% of the lakes in the 2012 survey less than 0.5 km2.
+
+PARAGRAPH about testing hypotheses
+
+In summary, the main goal of this work was to develop a simple yet robust lake photic temperature model for all lakes in the conterminous US while still capturing key drivers of near-surface lake temperature.
+Additionally, this work (something about open science and GitHub link).
 
 # Methods
+
+##Data
+We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment (NLA) (CITE).
+The NLA is a stratified random sample of lakes (great than 1 ha) across the United States.
+NLA sampling took place in 2007, 2012, and 2017.
+This research effort used the 2007 and 2012 sample years.
+The 2017 data are currently undergoing quality control before being released the public.
+For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).
+For each sampled lake, we used the mean temperature for all sampled depths of less than 2m (this being the depth typically considered the photic zone).
+
+Figure 1: Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes
+
+In addition to climate, there are additional factors clearly impacting lake temperature (e.g., surrounding land use, lake depth, size, and configuration, and elevation).
+To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages lmorpho and elevatr (CITE) as means to incorporate these types of data.
+Additionally, National Land Cover Database (NLCD) was our source for landcover data.
+Specifically, we calculated the percent impervious surface of a 3000m buffer for each lake.
+Lakes with partials buffers falling outside the US were excluded.
+
+As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).
+This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).
+PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).
+For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.
+
+To test the relative influence of both short and long-term temperature, we derived several measures for a lake’s local air temperature.
+Mean air temperatures for day of and the day before the sample date were extracted directly from the prism data.
+To understand longer term influences we calculated average mean air temperatures for periods 3, 7, and 30 days prior to the sample date.
+
+##Random Forest Modelling
+
+Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and measure of relative variable importance.
+Random forest modelling is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees).
+Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.
+All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
+Model performance is reported as mean square error and adjusted R.
+Additionally, we used percent increase of mean-squared error to assess variable importance.
+The percent increase in mean-squared error is a comparison between the true values of a variable and randomly permuted values of a specific variable.
 

### 2019-05-08 manuscript/manuscript.Rmd
Adding citation

@@
 During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].
-As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)(NASA GISS), Surface temperatures of lakes are also increasing globally (0.34 per decade from 1985 to 2009) [@o2015rapid].
+As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)[@hansen2010global], surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009) [@o2015rapid].
 The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.
-Ultimately, temperature changes will greatly impact all aspects of lake resource management.
+Ultimately, temperature changes will greatly impact every  aspects of lake resource management.
 For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].
 During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].
-As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over other phytoplankton and can proliferate quickly [@paerl2008blooms].
+As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].
 Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].
-Thus, our ability to understand and predict toxin cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.
-This need for accurate lake temperature forecasting is will become crucial for protecting human and environmental health.
+Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.
+This need for accurate lake temperature forecasting will become crucial for protecting human and environmental health.
 
@@
 Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).
-Such studies include: [@saeed2016water; @peeters2002modelin; @zhong2016recent].
+Such studies include: [@saeed2016water; @peeters2002modeling; @zhong2016recent].
 There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].

### 2019-06-06 R/figures.R
Building figures

@@ -1,0 +1,29 @@
+#' script to generate figures
+#' load up functions and packages
+#' 
+source(here::here("R/functions.R"))
+
+
+varsel_fig<-varsel_plot(var_sel_tmean)
+
+ggsave(here("figures/valselfig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
+       units = "in", dpi = 600)
+
+
+```{r all_var_sel_figure, eval=TRUE,results="asis",echo=FALSE, fig.cap="Variable selection plot for all variables.  Shows percent increase in mean squared error as a function of the number of variables.  \\label{fig:all_varsel_figure}",cache=FALSE}
+varsel_fig<-varsel_plot(var_sel_tmean)
+
+#Height maximum is 6.8 inches
+ggsave(filename = "ecosphere_final/hollisterES15-00703R_fig3.tiff",
+       plot = fig3,
+       dpi = 450,
+       #width = 6.1,
+       height = 6.8,
+       units = "in")
+x<-autocrop("ecosphere_final/hollisterES15-00703R_fig3.tiff",border = 5, 
+            outfile = "ecosphere_final/hollisterES15-00703R_fig3.tiff",
+            width = 6.1,
+            res = 450,
+            compression = "lzw")
+fig3
+```

### 2019-06-06 manuscript/manuscript.Rmd
Building figures
adding figures to manuscript

@@
 ```{r setup, include=FALSE, eval=FALSE}
-source(here::here("R/functions.R"), encoding = "UTF-8")
-opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
+# source(here::here("R/functions.R"), encoding = "UTF-8")
+library(captioner)
+#opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
 tab_num <- captioner(prefix = "Table")
@@
 options(scipen = 2, digits = 5)
+
 ```
@@
 
+![`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean‐squared error as a function of the number of variables.")`](../figures/varselfig.jpg)
+
 # Discussion and conclusions

### 2019-06-11 manuscript/manuscript.Rmd
Adding methods

@@
 
-Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and measure of relative variable importance.
-Random forest modelling is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees).
+Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.
+Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
 Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.
@@
 Model performance is reported as mean square error and adjusted R.
-Additionally, we used percent increase of mean-squared error to assess variable importance.
-The percent increase in mean-squared error is a comparison between the true values of a variable and randomly permuted values of a specific variable.
+
+Random forest does not require that users reduce the number of predictor variables; the random forest algorithm prevents overfitting and is not impacted by correlated predictor variables (Culter 2007).
+While it is unlikely that random forest models constructed with reduced numbers of predictor variables perform any better than models constructed with a full suite of available variables(Fox et al., 2017), reducing the number of our predictor variables was necessary in order to fulfill the long-term goals of this work.
+Several of our climatic predictor variables are computationally intensive to create for the entire conterminous United States.
+In order to use our final model for future forecasting or historical backcasting, we strove to create a robust predictive model while minimizing computational demands.
+To determine the optimal number and set of variables, we followed the variable selection method presented in (Hollister 2016).
+
+In addition to measure of overall model performance, we used percent increase of mean-squared error to assess variable importance.
+The percent increase in mean-squared error is a comparison between the true values of a variable and randomly permuted values of a specific variable on overall model performance.
 
 # Results
+
+We used 1185 data points from the 2007 NLA and 1097 from 2012 NLA.  Figure 1 illustrates the spatial distribution of the sampled lakes.
 

### 2019-06-11 R/functions.R

@@ -1,0 +1,116 @@
+# Installations
+pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
+          "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
+          "USAboundariesData","knitr","captioner")
+
+for(i in pkgs){
+  if(!i %in% installed.packages()){
+    install.packages(i, repos = c("http://packages.ropensci.org", 
+                                  "https://cran.rstudio.com"))
+  }
+}
+
+if(!requireNamespace("LAGOSNE")){
+  devtools::install_github("cont-limno/LAGOSNE")
+}
+
+# Library
+library(lubridate)
+library(broom)
+library(stringr)
+library(dplyr)
+library(readr)
+library(readxl)
+library(tidyr)
+library(here)
+library(ggplot2)
+library(hrbrthemes)
+library(Kendall)
+library(LAGOSNE)
+library(sf)
+library(USAboundaries)
+library(USAboundariesData)
+library(knitr)
+library(captioner)
+library(cowplot)
+
+#' Function to filter on number of years sampled per site
+#' @param df data frame to filter
+#' @param num_yrs minimum number of years
+filter_year <- function(df, num_yrs){
+  df %>% 
+    select(year,station_name) %>%
+    unique() %>%
+    group_by(station_name) %>%
+    tally() %>%
+    filter(n >= num_yrs) %>%
+    pull(station_name)
+}
+
+#' Function to generate figures
+#' @param df data frame with data to plot
+#' @param wqparam the water quality parameter to plot
+#' @param yvar the actual variable to plot, defaults to anomaly
+#' @param num_yrs minimum number of years needed to include a site
+#' @param write write out data used in the plot to a csv file
+#' @param title title for plot
+#' @param start_yr only plots years for which we have data for all params
+wq_trend_gg <- function(df, wqparam, 
+                        yvar = c("measurement_anmly","measurement_scale"), 
+                        num_yrs = 10, write = NULL, title = "",
+                        start_yr = 1993, error_bar = c("se", "sd"), ...){
+  yvar <- rlang::sym(match.arg(yvar))
+  error_bar <- rlang::sym(match.arg(error_bar))
+  
+  df1 <- df %>%
+    filter(param == wqparam) %>%
+    filter(year >= start_yr) %>%
+    filter(station_name %in% filter_year(., num_yrs))
+  
+  
+  df2 <- df1 %>%
[…]

### 2019-06-11 R/figures.R
Code for Figure 2

@@ -6,24 +6,7 @@ source(here::here("R/functions.R"))
 
 varsel_fig<-varsel_plot(var_sel_tmean)
 
-ggsave(here("figures/valselfig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
+ggsave(here::here("figures/valselfig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
 
 
-```{r all_var_sel_figure, eval=TRUE,results="asis",echo=FALSE, fig.cap="Variable selection plot for all variables.  Shows percent increase in mean squared error as a function of the number of variables.  \\label{fig:all_varsel_figure}",cache=FALSE}
-varsel_fig<-varsel_plot(var_sel_tmean)
-
-#Height maximum is 6.8 inches
-ggsave(filename = "ecosphere_final/hollisterES15-00703R_fig3.tiff",
-       plot = fig3,
-       dpi = 450,
-       #width = 6.1,
-       height = 6.8,
-       units = "in")
-x<-autocrop("ecosphere_final/hollisterES15-00703R_fig3.tiff",border = 5, 
-            outfile = "ecosphere_final/hollisterES15-00703R_fig3.tiff",
-            width = 6.1,
-            res = 450,
-            compression = "lzw")
-fig3
-```

### 2019-06-11 manuscript/manuscript.Rmd
small edits to results

@@
 
+Figure 1
+
+Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.
+The variable selection process identified a reduced model with 7 variables (Fig 2).
+The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.
+
 ![`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean‐squared error as a function of the number of variables.")`](../figures/varselfig.jpg)
+
+The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R^2 of 0.88.
+The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (Figure 3).
+The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (Figure 4).
 

### 2019-07-30 manuscript/manuscript.Rmd
edits to figure and model code

@@
 
+Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States.
+Average ambient air temperature from the sample date, data obtained from the PrRISM Climate Group, was the most important variable imapcting the final model's accuracy.
+
 # Bibliography

### 2019-07-31 manuscript/manuscript.Rmd
updated figure caption for varsel fig and fixed R cite in .bib

@@
 
-```{r setup, include=FALSE, eval=FALSE}
+```{r setup, include=FALSE, eval=TRUE}
 # source(here::here("R/functions.R"), encoding = "UTF-8")
@@
 options(scipen = 2, digits = 5)
-
 ```
@@
 
-##Random Forest Modelling
+## Random Forest Modelling
 
@@
 Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.
-The variable selection process identified a reduced model with 7 variables (Fig 2).
+The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables.
+Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
 The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.
 
-![`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean‐squared error as a function of the number of variables.")`](../figures/varselfig.jpg)
+![`r fig_num("varsel_fig")`](../figures/valselfig.jpg)
 

### 2019-08-01 R/model.Rmd
B: saved a copy of RFAll to a local directory
added line to save rds
edits to figure and model code
Added rough draft of figures to the manuscript

@@
               max_depth, volume, elevation)
+
+nla.all<-rename(nla.all,Date=date,Average.Temperature=tmean,Longitude=longitude,
+                 Thirty.Day.Average.Temperature=tmean_avg30,Elevation=elevation,
+                 Latitude=latitude,Surface.Area=surface_area,
+                 Shoreline.Length=shoreline_length)
 
@@
 
-varImpPlot(RFAll)
+varImpPlot(RFAll,type=1)
+
+# B: I'm saving a copy of the model for later use.  It is too big to github so i added it to .gitignore
+saveRDS(RFAll, here::here('data/RFAll20190503.Rds'))
+```
 
 ```
-
-mean_run <- nla.cc %>%
-
-  select(tmean_2m,var_sel_tmean$vars[[9]]) %>%
-
-  randomForest(tmean_2m ~ ., data = ., importance = TRUE,
-
-               ntree = 1000)
-
-mean_run
-
-varImpPlot(mean_run)
-
-```

### 2019-08-01 manuscript/manuscript.Rmd
Added rough draft of figures to the manuscript

@@
 
-![`r fig_num("varsel_fig")`](../figures/valselfig.jpg)
+![`r fig_num("varsel_fig")`](../figures/varselfig.jpg)
 
 The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R^2 of 0.88.
-The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (Figure 3).
-The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (Figure 4).
+The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable selection plot for all variables.
+Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
+The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Variable selection plot for all variables.
+Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
+
+![`r fig_num("varImpPlot")`](../figures/varImpPlot.jpg)
+
+![`r fig_num("partPlot")`](../figures/partPlot.jpg)
 

### 2019-08-01 R/figures.R

@@ -3,10 +3,35 @@
 #' 
 source(here::here("R/functions.R"))
 
-
+###Variable selection figure 
 varsel_fig<-varsel_plot(var_sel_tmean)
 
-ggsave(here::here("figures/valselfig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
+ggsave(here::here("figures/varSelFig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
+
+
+###Variable importance figure
+
+jpeg("figures/varImpPlot.jpg")
+varImpPlot(RFAll,type=1,main="")
+dev.off()
+
+
+
+###Partial Dependence Plots 
+jpeg("figures/partPlot.jpg")
+par(mfrow=c(4,2))
+
+partialPlot(RFAll, nla.all,Date,main="")
+partialPlot(RFAll, nla.all,Average.Temperature,main="")
+partialPlot(RFAll, nla.all,Longitude,main="")
+partialPlot(RFAll, nla.all,Thirty.Day.Average.Temperature,main="")
+partialPlot(RFAll, nla.all,Elevation,main="")
+partialPlot(RFAll, nla.all,Latitude,main="")
+partialPlot(RFAll, nla.all,Surface.Area,main="")
+partialPlot(RFAll, nla.all,Shoreline.Length,main="")
+dev.off()
+
+
 
 

### 2019-08-01 R/varImportance.R

@@ -1,0 +1,20 @@
+#' Variable Importance Summary
+#' 
+#' @param rf_imp a Random Forest Importance Object
+#' @param label A label for the model
+#' @export
+varImportance<-function(rf_imp,label=NULL){
+  
+  if(!is.null(label)){
+    xdf<-data.frame(model_id=rep(label,nrow(rf_imp)),
+                    variable_names=row.names(rf_imp),
+                    mean_decrease_acc=rf_imp[,1],
+                    mean_decrease_gini=rf_imp[,2])
+  } else {
+    xdf<-data.frame(model_id=rep(NA,nrow(rf_imp)),
+                    variable_names=row.names(rf_imp),
+                    mean_decrease_acc=rf_imp[,1],
+                    mean_decrease_gini=rf_imp[,2])
+  }   
+  return(xdf)
+}

### 2019-08-01 R/importancePlot.R
More figure updates

@@ -1,0 +1,68 @@
+#' Random Forest Variable Importance Plotting
+#' 
+#' @param rf a randomForest object
+#' @param type the type of importance to plot "acc" for Mean Decrease Accuracy, and 
+#'              "gini for Mean Decrease Gini.
+#' @param ... any items that can be passed to geom_point.  Mostly used to control size
+#'            and color of points.
+#' 
+#' @export
+#' @import ggplot2
+#' @examples
+#' require(wesanderson)
+#' data(iris)
+#' iris_rf<-randomForest(iris[,1:4],iris[,5],importance=T,proximity=T)
+#' importancePlot(iris_rf,'acc',size=10,aes(colour=wes.palette(4, 'GrandBudapest')))
+#' importancePlot(iris_rf,'gini',size=10,aes(colour=wes.palette(4, 'GrandBudapest')))
+
+importancePlot <- function(rf, sumtbl=NULL, data_def = NULL, type = c("acc", "gini"), ...) {
+  
+  idf<-importance(rf)
+  # set up data.frame for ggplot
+  if (type == "gini") {
+    imp_df <- data.frame(variables = names(idf[,dim(idf)[2]]), 
+                         Mean_Decrease = idf[,dim(idf)[2]])
+    o <- order(imp_df$Mean_Decrease, decreasing = FALSE)
+    imp_df$variables <- factor(imp_df$variables, 
+                               levels = imp_df$variables[o], ordered = T)
+    label <- "Mean Decrease Gini"
+  } else if (type == "acc") {
+    imp_df <- data.frame(variables = names(idf[,dim(idf)[2]-1]), 
+                         Mean_Decrease = idf[,dim(idf)[2]-1])
+    o <- order(imp_df$Mean_Decrease, decreasing = FALSE)
+    imp_df$variables <- factor(imp_df$variables, 
+                               levels = imp_df$variables[o], ordered = T)
+    label <- "Percent Increase in Mean Square Error"
+  } else {
+    stop("Type not specificed correctly")
+  }
+  if(!is.null(sumtbl)){
+    imp_df<-merge(imp_df,sumtbl,by.x="variables",by.y="Variable")
+  } 
+  if (!is.null(data_def)){
+    imp_df$variables<-data_def$description[match(imp_df$variables,data_def$variable_names)]
+    o <- order(imp_df$Mean_Decrease, decreasing = FALSE)
+    imp_df$variables <- factor(imp_df$variables, 
[…]

### 2019-08-08 manuscript/manuscript_old.Rmd
trying to fix something

--- 2019-08-01 manuscript/manuscript.Rmd
@@
 Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.
-This need for accurate lake temperature forecasting will become crucial for protecting human and environmental health.
+This need for accurate lake temperature forecasting will be crucial for protecting human and environmental health.
 
 Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.
-These studies typically vary in number of lakes studied, complexity of modelling approach, and time interval.
+These studies typically vary in number of lakes studied, complexity of modelling approach, and study period.
 Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).
@@
 The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
-The NLA is stratified random sample of all lakes in the conterminous United State.
-The NLA was conducted in 2007 and then repeated in 2012.
-Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.
-The uniqueness of this data set allows us to build a robust lake photic zone temperature model for all US lakes.
-
-The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
 The NLA is stratified random sample of all lakes in the conterminous United States.
@@
 Additionally, small lakes were well represented with more than 50% of the lakes in the 2012 survey less than 0.5 km2.
-
-PARAGRAPH about testing hypotheses
 
@@
 Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
-Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.
-All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
+Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.(ADD SENTENCE)  All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
 Model performance is reported as mean square error and adjusted R.
@@
 Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.
-The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables.
+The variable selection process identified a reduced model with eight variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables.
 Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
@@
 Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States.
-Average ambient air temperature from the sample date, data obtained from the PrRISM Climate Group, was the most important variable imapcting the final model's accuracy.
+The final model had a mean square error of 2.19 and an adjusted R^2 of 0.88.
+The sampling date and that day’s average ambient air temperature, data obtained from the PRISM Climate Group, were the most important variables impacting the final model’s accuracy.
+Assuming that ambient air temperature date is available, this model allows us to predict lake photic zone temperature (Importance of this ability??).
+
+For the final model, average ambient air temperature and the average temperature of the prior 30 days were selected.
+Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.
+It is likely that the three and seven day averages are not providing the model with unique information.
+Typical the three and seven day averages are fairly close the day-of average.
+Ambient air temperature does not change drastically over such a short time scale.
+Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.
+Yet the longer term 30-day average does have an impact on lake photic zone temperature.
+The 30-day average provides us with this information about the temperature intensity leading up to the sample date.
+Not just how far along into summer; that information would be captured in the date parameter.
+But more specifically, this measures the long term thermal heating happening at a site.
+
+In addition to the several average air temperature variables, we included land-use/land cover variables.
+Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this measured).
+This variables were included based on the assumption that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
+Yet neither of these variables were selected in the final model.
+Even though the land-use variables were not selected that does not mean development and impervious were not impactful.
+This urban-heat effect on lakes may have been adequately captured in the average ambient air temperature.
+Therefore, making the land-use variables redundant.
+Regardless these variables did not independently contribute to the model’s accuracy.
 

### 2019-08-08 manuscript/manuscript.Rmd
fixing merge conflict

@@
 The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables.
-Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
+Shows percent increase in mean square error as a function of the number of variables.", display = "cite")`).
 The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.
@@
 The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R^2 of 0.88.
-The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable selection plot for all variables.
-Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
-The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Variable selection plot for all variables.
-Shows percent increase in mean‐squared error as a function of the number of variables.", display = "cite")`).
+The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.
+Shows precent increase in mean square error.
+Higher values indicate a higher impact on overall model accuracy.", display = "cite")`).
+The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Partial dependence plots for selected variables.", display = "cite")`).
 

### 2019-08-09 manuscript/manuscript.Rmd
Added paragraph to discussion

@@
 In addition to the several average air temperature variables, we included land-use/land cover variables.
-Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this measured).
+Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this is measured).
 This variables were included based on the assumption that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
@@
 
+Insert paragraph on model shortcomings
+
+An advantage of this model is that we are now able to forecast and backcast lake photic zone temperature.
+While there are numerous lakes that have substantial historical temperature data, these data are not consistently available for large regions.
+This inhibits our ability to make large scale predictions about changes to lake dynamics (e.g., stratification and mixing) and ecological risk (e.g., anticipated changes in cyanobacteria bloom frequency).
+At the continental scale, essentially, we could identify area that might have a higher ecological risk.
+Yet if there was an individual lake of concern that lack temperature, this model would be as effective.
+
+Final paragraph
+
 # Bibliography

### 2019-08-12 manuscript/manuscript.Rmd
Adding to discussion

@@
 
-In addition to the several average air temperature variables, we included land-use/land cover variables.
+In addition to the several derived air temperature variables, we included land-use/land cover variables in our initial variable selection process.
 Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this is measured).
-This variables were included based on the assumption that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
+This variables were included based on the hypothesis that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
 Yet neither of these variables were selected in the final model.
-Even though the land-use variables were not selected that does not mean development and impervious were not impactful.
+Even though the land-use variables were not selected that does not mean development and impervious are not impactful.
 This urban-heat effect on lakes may have been adequately captured in the average ambient air temperature.

### 2019-08-26 manuscript/manuscript.Rmd
Added Stephen and Bryan's edits

@@
 Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States.
-The final model had a mean square error of 2.19 and an adjusted R^2 of 0.88.
-The sampling date and that day’s average ambient air temperature, data obtained from the PRISM Climate Group, were the most important variables impacting the final model’s accuracy.
-Assuming that ambient air temperature date is available, this model allows us to predict lake photic zone temperature (Importance of this ability??).
+The final model has a mean square error of 2.19 and an adjusted R^2 of 0.88.
+The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
+Given the importance of temperature to a lake ecosystem, especially cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers.
+Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.
 
@@
 
-An advantage of this model is that we are now able to forecast and backcast lake photic zone temperature.
-While there are numerous lakes that have substantial historical temperature data, these data are not consistently available for large regions.
-This inhibits our ability to make large scale predictions about changes to lake dynamics (e.g., stratification and mixing) and ecological risk (e.g., anticipated changes in cyanobacteria bloom frequency).
-At the continental scale, essentially, we could identify area that might have a higher ecological risk.
-Yet if there was an individual lake of concern that lack temperature, this model would be as effective.
-
-Final paragraph
+Despite being one of the most common measurements collected by limnologists, lake temperature datasets that cover long periods of time are very difficult to obtain.
+Sharma et al (2015) have compiled summer lake temperature data for 291 lakes for the period 1985-2009.
+This may be the largest lake temperature database to date, however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.
+One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.
+The model we present has proven to be accurate and will allow us to backcast lake temperatures for all the > 300,000 lakes included in NHDplus for the period of time covered by the PRISM climate predictions (1981 to present).
+This dataset will allow us to investigate how photic zone temperatures vary both spatially and temporally across the United States.
+This database is being developed and, when complete, will be made available as an open source data set.
 

### 2019-09-05 manuscript/manuscript.Rmd
Edits mainly to Intro

@@
 ---
-title: "Lake Photic Zone Temperature accross the Conterminous United States"
+title: "Lake Photic Zone Temperature across the Conterminous United States"
 author:
@@
 
-Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.
-These studies typically vary in number of lakes studied, complexity of modelling approach, and study period.
+Because to the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.
+Models typically vary in number of lakes studied, complexity of modelling approach, and study interval.
 Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).
@@
 By and large, air temperature and lake size are often the only selected predictor variables.
-Finally, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.
+On the whole, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.
 Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].
@@
 
-Therefore, we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
+In summary, we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
 Modelling lake temperature requires a large amount of data, and therefore study lakes are often selected opportunistically, which may introduce a spatial bias.
@@
 
-The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
-The NLA is stratified random sample of all lakes in the conterminous United States.
-The NLA was conducted in 2007 and then repeated in 2012.
+The presented modelling effort uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
+The NLA is a stratified random sample of all lakes in the conterminous United States repeated every five years beginning in 2007.
 Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.

### 2019-09-05 manuscript/manuscript.Rmd
small edits to MS

@@
 
-In summary, the main goal of this work was to develop a simple yet robust lake photic temperature model for all lakes in the conterminous US while still capturing key drivers of near-surface lake temperature.
+To summarize, the main goal of this work was to develop a simple yet robust lake photic temperature model for all lakes in the conterminous US that capture key drivers of near-surface lake temperature.
 Additionally, this work (something about open science and GitHub link).

### 2019-09-19 manuscript/manuscript.Rmd
jeff going overboard with edits

@@
 
-```{r analysis, include=FALSE, eval=FALSE}
-#got to figure out path stuff before render.
-source(here::here("R/data_available.R"), encoding = "UTF-8") 
-source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
+```{r model, include=FALSE, eval=TRUE, cache = TRUE}
+source(here::here("R/model.R"), encoding = "UTF-8")
+```
+```{r figures, include=FALSE}
 source(here::here("R/figures.R"), encoding = "UTF-8")
@@
 In addition to climate, there are additional factors clearly impacting lake temperature (e.g., surrounding land use, lake depth, size, and configuration, and elevation).
-To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages lmorpho and elevatr (CITE) as means to incorporate these types of data.
+To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages `lakemorpho` to calculate a suite of lake morphometry metric and `elevatr` to access digital elevation models for each lake [@hollister2017lakemorpho; @hollister2017elevatr].
 Additionally, National Land Cover Database (NLCD) was our source for landcover data.

### 2019-09-19 R/figures.R

@@ -1,7 +1,12 @@
 #' script to generate figures
-#' load up functions and packages
 #' 
-source(here::here("R/functions.R"))
+source(here::here("R/varsel_plot.R"))
+source(here::here("R/varimp_plot.R"))
+source(here::here("R/partial_plot.R"))
+
+extrafont::loadfonts(device="win")
+library(ggplot2)
+library(hrbrthemes)
 
 ###Variable selection figure 
 varsel_fig<-varsel_plot(var_sel_tmean)

### 2019-09-19 R/functions.R

@@ -10,10 +10,6 @@ for(i in pkgs){
   }
 }
 
-if(!requireNamespace("LAGOSNE")){
-  devtools::install_github("cont-limno/LAGOSNE")
-}
-
 # Library
 library(lubridate)
 library(broom)
@@ -34,83 +30,27 @@ library(knitr)
 library(captioner)
 library(cowplot)
 
-#' Function to filter on number of years sampled per site
-#' @param df data frame to filter
-#' @param num_yrs minimum number of years
-filter_year <- function(df, num_yrs){
-  df %>% 
-    select(year,station_name) %>%
-    unique() %>%
-    group_by(station_name) %>%
-    tally() %>%
-    filter(n >= num_yrs) %>%
-    pull(station_name)
+# https://raw.githubusercontent.com/USEPA/LakeTrophicModelling/master/R/varsel_plot.R
+
+#' Plot varsel_regression output
+#' 
+#' @param vs_reg_obj Object from output of \link{varsel_regression_rf}.
+#' @export
+#' @import ggplot2
+varsel_plot <- function(vs_reg_obj){
+  dat <- data.frame(num = vs_reg_obj$num_var, mse = vs_reg_obj$mse)
+  browser()
+  plot_out<-ggplot(dat, aes(x=num,y=mse))+geom_point(size=2.5)+
+    theme(text = element_text(family="sans"),
+          panel.background = element_blank(), #panel.grid = element_blank(), 
+          panel.border = element_rect(fill = NA), 
+          plot.title  = element_text(family="sans",size=15,face="bold",vjust=1.1),
+          legend.key = element_rect(fill = 'white'),
+          legend.text = element_text(family="sans",size=15), legend.title = element_text(size=15),
+          axis.title.x = element_text(family="sans",vjust = -0.5, size = 12),
+          axis.title.y = element_text(family="sans",vjust = 1.5, size = 12),
+          axis.text.x = element_text(family="sans",size = 11),
+          axis.text.y = element_text(family="sans",size = 11))+
+    labs(x="Number of variables in model", y="Root Mean Square Error")
+  return(plot_out)
 }
-
-#' Function to generate figures
-#' @param df data frame with data to plot
-#' @param wqparam the water quality parameter to plot
-#' @param yvar the actual variable to plot, defaults to anomaly
-#' @param num_yrs minimum number of years needed to include a site
-#' @param write write out data used in the plot to a csv file
-#' @param title title for plot
[…]

### 2019-09-19 R/model.R

@@ -1,0 +1,48 @@
+# Library
+library(randomForest)
+library(dplyr)
+library(purrr)
+library(here)
+library(broom)
+library(lubridate)
+
+source(here::here('R/varsel_regression_rf.R'))
+
+##Reading in and creating data sets
+nla<-read.csv(here::here('data/nla_base.csv'))
+x<-complete.cases(nla)
+nla.cc<-nla[x,]
+nla.cc$year<-factor(nla.cc$year)
+nla.cc$date <- lubridate::yday(lubridate::ymd(nla.cc$date))
+
+nla.all <-  filter(nla.cc, tmean_2m < 70) %>%
+              dplyr::select(tmean_2m, year,date, longitude, latitude, tmean,
+              tmean_avg30,percent_impervious,surface_area,shoreline_length,
+              shoreline_dev, max_length, max_width, mean_width, max_depth,
+              max_depth, volume, elevation)
+
+nla.all<-rename(nla.all,Date=date,Average.Temperature=tmean,Longitude=longitude,
+                 Thirty.Day.Average.Temperature=tmean_avg30,Elevation=elevation,
+                 Latitude=latitude,Surface.Area=surface_area,
+                 Shoreline.Length=shoreline_length)
+
+##Variable selection for 2007&2012 model
+set.seed(42)
+var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
+
+var_sel_tmean$vars[[7]]
+
+set.seed(42)
+RFAll <- nla.all %>%
+
+  select(tmean_2m,var_sel_tmean$vars[[7]]) %>%
+
+  randomForest(tmean_2m ~ ., data = ., importance = TRUE,
+
+               ntree = 10000)
+
+
+
+
+
+

### 2019-09-19 R/varsel_plot.R

@@ -7,17 +7,24 @@
 #' @import ggplot2
 varsel_plot <- function(vs_reg_obj){
   dat <- data.frame(num = vs_reg_obj$num_var, mse = vs_reg_obj$mse)
-  plot_out<-ggplot(dat, aes(x=num,y=mse))+geom_point(size=2.5)+
-    theme(text = element_text(family="sans"),
-          panel.background = element_blank(), #panel.grid = element_blank(), 
-          panel.border = element_rect(fill = NA), 
-          plot.title  = element_text(family="sans",size=15,face="bold",vjust=1.1),
-          legend.key = element_rect(fill = 'white'),
-          legend.text = element_text(family="sans",size=15), legend.title = element_text(size=15),
-          axis.title.x = element_text(family="sans",vjust = -0.5, size = 12),
-          axis.title.y = element_text(family="sans",vjust = 1.5, size = 12),
-          axis.text.x = element_text(family="sans",size = 11),
-          axis.text.y = element_text(family="sans",size = 11))+
+  #browser()
+  suppressWarnings({
+  plot_out<-ggplot(dat, aes(x=num,y=mse)) +
+    geom_point(size=2.5) +
+    theme_ipsum() +
     labs(x="Number of variables in model", y="Root Mean Square Error")
-  return(plot_out)
+  plot_out
+  })
 }
+
+
+#theme(text = element_text(family="sans"),
+#      panel.background = element_blank(), #panel.grid = element_blank(), 
+#      panel.border = element_rect(fill = NA), 
+#      plot.title  = element_text(family="sans",size=15,face="bold",vjust=1.1),
+#      legend.key = element_rect(fill = 'white'),
+#      legend.text = element_text(family="sans",size=15), legend.title = element_text(size=15),
+#      axis.title.x = element_text(family="sans",vjust = -0.5, size = 12),
+#      axis.title.y = element_text(family="sans",vjust = 1.5, size = 12),
+#      axis.text.x = element_text(family="sans",size = 11),
+#     axis.text.y = element_text(family="sans",size = 11))+

### 2019-09-20 R/figures.R
starting on figures

@@ -4,13 +4,13 @@ source(here::here("R/varsel_plot.R"))
 source(here::here("R/varimp_plot.R"))
 source(here::here("R/partial_plot.R"))
 
-extrafont::loadfonts(device="win")
+#extrafont::loadfonts(device="win")
 library(ggplot2)
 library(hrbrthemes)
 
 ###Variable selection figure 
 varsel_fig<-varsel_plot(var_sel_tmean)
-
+varsel_fig
 ggsave(here::here("figures/varSelFig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
 

### 2019-09-20 R/model.R

@@ -30,8 +30,6 @@ nla.all<-rename(nla.all,Date=date,Average.Temperature=tmean,Longitude=longitude,
 set.seed(42)
 var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
 
-var_sel_tmean$vars[[7]]
-
 set.seed(42)
 RFAll <- nla.all %>%
 

### 2019-09-20 R/varsel_plot.R

@@ -8,13 +8,11 @@
 varsel_plot <- function(vs_reg_obj){
   dat <- data.frame(num = vs_reg_obj$num_var, mse = vs_reg_obj$mse)
   #browser()
-  suppressWarnings({
   plot_out<-ggplot(dat, aes(x=num,y=mse)) +
-    geom_point(size=2.5) +
-    theme_ipsum() +
+    geom_point(size = 3) +
+    theme_ipsum_rc(base_size = 12, axis_title_size = 12) +
     labs(x="Number of variables in model", y="Root Mean Square Error")
   plot_out
-  })
 }
 
 

### 2019-09-20 R/figures.R
progres on figures.  Need to calculate partial dependencies.

@@ -10,16 +10,18 @@ library(hrbrthemes)
 
 ###Variable selection figure 
 varsel_fig<-varsel_plot(var_sel_tmean)
-varsel_fig
-ggsave(here::here("figures/varSelFig.jpg"), varsel_fig, width = 7.5, height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here::here("figures/varSelFig.jpg"), varsel_fig, width = 7.5, 
+       height = 5.625, units = "in", dpi = 600)
 
 
 ###Variable importance figure
-
-jpeg("figures/varImpPlot.jpg")
-varImpPlot(RFAll,type=1,main="")
-dev.off()
+variable_names <- c("Avg. temperature", "Date", "Longitude", 
+                    "30-day avg. temperature", "Elevation", "Latitude", 
+                    "Lake shoreline length", "Lake area")
+varimp_fig <- varimp_plot(RFAll, variable_names)
+varimp_fig
+ggsave(here::here("figures/varImpPlot.jpg"), varimp_fig, width = 7.5, 
+       height = 5.625, units = "in", dpi = 600)
 
 
 

### 2019-09-20 R/model.R

@@ -30,16 +30,13 @@ nla.all<-rename(nla.all,Date=date,Average.Temperature=tmean,Longitude=longitude,
 set.seed(42)
 var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
 
+
+nla_select <- select(nla.all, tmean_2m,var_sel_tmean$vars[[7]])
 set.seed(42)
-RFAll <- nla.all %>%
-
-  select(tmean_2m,var_sel_tmean$vars[[7]]) %>%
-
-  randomForest(tmean_2m ~ ., data = ., importance = TRUE,
-
+RFAll <- randomForest(tmean_2m ~ ., data = nla_select, importance = TRUE,
                ntree = 10000)
 
-
+## Calculate partial dependencies
 
 
 

### 2019-09-20 R/partial_plot.R

@@ -1,0 +1,14 @@
+#' Random Forest Partial Dependency Plots
+#' 
+#' This function creates the partial dependency plots used in the Kreakie et al. 
+#' paper on random forest modeling of photic zone temperature
+#' 
+#' @param rfobj an random forest object. see \link{\code{randomForest}} for 
+#'              details on creating this object
+#' @param dataobj data used to calculate partial dependency
+#' @param variable variable to plot
+#'              
+partial_plot <- function(rfobj, dataobj, variable){
+  browser()
+  
+}

### 2019-09-20 R/varimp_plot.R

@@ -1,0 +1,27 @@
+#' Random Forest Variable Importance Plots
+#' 
+#' This function creates the variable importance plots used in the Kreakie et al. 
+#' paper on random forest modeling of photic zone temperature
+#' 
+#' @param rfobj A random forest object. see \link{\code{randomForest}} for 
+#'              details on creating this object.
+#' @param var_names Optional variable names for the y axis.  Must be in same 
+#'                  order as row.names(rfobj$importance).
+#'              
+varimp_plot <- function(rfobj, var_names = NULL){
+  if(is.null(var_names)){var_names <- row.names(rfobj$importance)}
+  perc_inc_mse <- data.frame(rfobj$importance)$X.IncMSE/rfobj$importanceSD
+  var_names <- reorder(factor(var_names), perc_inc_mse)
+  varimp_df <- data.frame(var_names, perc_inc_mse, stringsAsFactors = FALSE)
+  plot_out <- ggplot(varimp_df, aes(y = perc_inc_mse, x = var_names)) +
+    geom_bar(stat = "identity", width = 0.5) +
+    coord_flip() +
+    labs(x = "Independent variables", 
+         y = "Percent increase mean square error") +
+    theme_ipsum_rc(base_size = 12, axis_title_size = 12) +
+    theme(axis.title.y = element_text(margin = margin(t = 0, r = 0.3, b = 0, 
+                                                      l = 0, "cm")),
+          axis.title.x = element_text(margin = margin(t = 0.3, r = 0, b = 0, 
+                                                      l = 0, "cm")))
+  plot_out
+}

### 2019-09-20 R/varsel_plot.R

@@ -11,18 +11,10 @@ varsel_plot <- function(vs_reg_obj){
   plot_out<-ggplot(dat, aes(x=num,y=mse)) +
     geom_point(size = 3) +
     theme_ipsum_rc(base_size = 12, axis_title_size = 12) +
-    labs(x="Number of variables in model", y="Root Mean Square Error")
+    labs(x="Number of variables in model", y="Root Mean Square Error") +
+    theme(axis.title.y = element_text(margin = margin(t = 0, r = 0.3, b = 0, 
+                                                      l = 0, "cm")),
+          axis.title.x = element_text(margin = margin(t = 0.3, r = 0, b = 0, 
+                                                      l = 0, "cm")))
   plot_out
 }
-
-
-#theme(text = element_text(family="sans"),
-#      panel.background = element_blank(), #panel.grid = element_blank(), 
-#      panel.border = element_rect(fill = NA), 
-#      plot.title  = element_text(family="sans",size=15,face="bold",vjust=1.1),
-#      legend.key = element_rect(fill = 'white'),
-#      legend.text = element_text(family="sans",size=15), legend.title = element_text(size=15),
-#      axis.title.x = element_text(family="sans",vjust = -0.5, size = 12),
-#      axis.title.y = element_text(family="sans",vjust = 1.5, size = 12),
-#      axis.text.x = element_text(family="sans",size = 11),
-#     axis.text.y = element_text(family="sans",size = 11))+

### 2019-10-28 R/figures.R
started on parital plot functions

@@ -26,18 +26,15 @@ ggsave(here::here("figures/varImpPlot.jpg"), varimp_fig, width = 7.5,
 
 
 ###Partial Dependence Plots 
-jpeg("figures/partPlot.jpg")
-par(mfrow=c(4,2))
+partplot_avg_temp <- partialPlot(RFAll, nla.all, Average.Temperature, plot = FALSE)
+partplot_long <- partialPlot(RFAll, nla.all,Longitude, plot = FALSE)
+partplot_thirty_day <- partialPlot(RFAll, nla.all,Thirty.Day.Average.Temperature, plot = FALSE)
+partplot_elev <- partialPlot(RFAll, nla.all,Elevation, plot = FALSE)
+partplot_lat <- partialPlot(RFAll, nla.all,Latitude, plot = FALSE)
+partplot_surf_area <- partialPlot(RFAll, nla.all,Surface.Area, plot = FALSE)
+partplot_shoreline_leng <- partialPlot(RFAll, nla.all,Shoreline.Length, plot = FALSE)
 
-partialPlot(RFAll, nla.all,Date,main="")
-partialPlot(RFAll, nla.all,Average.Temperature,main="")
-partialPlot(RFAll, nla.all,Longitude,main="")
-partialPlot(RFAll, nla.all,Thirty.Day.Average.Temperature,main="")
-partialPlot(RFAll, nla.all,Elevation,main="")
-partialPlot(RFAll, nla.all,Latitude,main="")
-partialPlot(RFAll, nla.all,Surface.Area,main="")
-partialPlot(RFAll, nla.all,Shoreline.Length,main="")
-dev.off()
+
 
 
 

### 2019-10-28 R/partial_plot.R

@@ -3,12 +3,9 @@
 #' This function creates the partial dependency plots used in the Kreakie et al. 
 #' paper on random forest modeling of photic zone temperature
 #' 
-#' @param rfobj an random forest object. see \link{\code{randomForest}} for 
-#'              details on creating this object
-#' @param dataobj data used to calculate partial dependency
-#' @param variable variable to plot
+#' @param partial_data an data frame with partial dependency calcs and variable names
 #'              
-partial_plot <- function(rfobj, dataobj, variable){
-  browser()
-  
+partial_plot <- function(partial_data){
+
+
 }

### 2019-10-28 R/varimp_plot.R

@@ -19,9 +19,9 @@ varimp_plot <- function(rfobj, var_names = NULL){
     labs(x = "Independent variables", 
          y = "Percent increase mean square error") +
     theme_ipsum_rc(base_size = 12, axis_title_size = 12) +
-    theme(axis.title.y = element_text(margin = margin(t = 0, r = 0.3, b = 0, 
+    theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 0.3, b = 0, 
                                                       l = 0, "cm")),
-          axis.title.x = element_text(margin = margin(t = 0.3, r = 0, b = 0, 
+          axis.title.x = element_text(margin = ggplot2::margin(t = 0.3, r = 0, b = 0, 
                                                       l = 0, "cm")))
   plot_out
 }

### 2019-10-28 manuscript/manuscript.Rmd

--- 2019-09-05 manuscript/manuscript.Rmd
@@
 
+Using only easily obtainable metrics to estimate lake temperature was a primary focus in the creation of this model.
+Several studies (e.g.
+[@piccolroaz2018predictability]) have used similar metrics with good results (RMSE=0.5 °C) to estimate lake temperature but were applied to a single, or a small number of lakes.
+Other studies have used more complex models (e.g.
+GLM) to yield good results (RMSE = 1.62 °C for the epilimnion) at varying scales [@bruce2018multi; @hipsey2019general].
+However, these models require more complex input variables which are not available for all lakes and may be biased towards large lakes.
+Using the NLA data and simple metrics allowed this model to be applicable to any lake within the conterminous United States using minimal input data with a small increase in RMSE.
+Additionally, this model is predictive and the model and data will be made publicly available.
+
 For the final model, average ambient air temperature and the average temperature of the prior 30 days were selected.

### 2020-06-26 R/model.Rmd
updated location of rf model

@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 # B: I'm saving a copy of the model for later use.  It is too big to github so i added it to .gitignore
-saveRDS(RFAll, here::here('data/RFAll20190503.Rds'))
+saveRDS(RFAll, "L:/Public/Milstead_Lakes/prism/RFAll20190503.Rds")
 ```

### 2020-06-27 manuscript/manuscript.Rmd
merged Betty's edits, and ignored manuscript cache.

@@
 author:
-- Kreakie, B. J. ^\*^ ^*1*^, Milstead, W. Bryan. ^*1*^, Hollister. J. W. ^*1*^, Shivers,
-  S. ^*2*^, 
+- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead, W. Bryan. ^*1*^,
 - ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
@@
 
-```{r model, include=FALSE, eval=TRUE, cache = TRUE}
+```{r model, include=FALSE, eval=TRUE, cache=TRUE}
 source(here::here("R/model.R"), encoding = "UTF-8")
 ```
-```{r figures, include=FALSE}
+```{r figures, include=FALSE, cache=TRUE}
 source(here::here("R/figures.R"), encoding = "UTF-8")
@@
 
+We included numerous predictor variables that are hypothesized to impact lake photic zone.
 As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).
@@
 
-Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States.
+Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States.
 The final model has a mean square error of 2.19 and an adjusted R^2 of 0.88.
 The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
-Given the importance of temperature to a lake ecosystem, especially cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers.
+Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers.
 Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.
@@
 
-For the final model, average ambient air temperature and the average temperature of the prior 30 days were selected.
+The final model included, average ambient air temperature and the average temperature of the prior 30 days.
 Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.
@@
 In addition to the several derived air temperature variables, we included land-use/land cover variables in our initial variable selection process.
-Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this is measured).
-This variables were included based on the hypothesis that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
+Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development.
+These variables were included based on the hypothesis that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
 Yet neither of these variables were selected in the final model.
@@
 Regardless these variables did not independently contribute to the model’s accuracy.
-
-Insert paragraph on model shortcomings
 

### 2020-06-27 R/figures.R

@@ -3,7 +3,7 @@
 source(here::here("R/varsel_plot.R"))
 source(here::here("R/varimp_plot.R"))
 source(here::here("R/partial_plot.R"))
-source(here::here("R/model.R"))
+#source(here::here("R/model.R"))
 
 #extrafont::loadfonts(device="win")
 library(ggplot2)
@@ -12,7 +12,7 @@ library(readr)
 
 ###Variable selection figure 
 varsel_fig<-varsel_plot(var_sel_tmean)
-varsel_fig
+#varsel_fig
 ggsave(here::here("figures/varSelFig.jpg"), varsel_fig, width = 7.5, 
        height = 5.625, units = "in", dpi = 600)
 
@@ -22,7 +22,7 @@ variable_names <- c("Avg. temperature", "Date", "Longitude",
                     "30-day avg. temperature", "Elevation", "Latitude", 
                     "Lake shoreline length", "Lake area")
 varimp_fig <- varimp_plot(RFAll, variable_names)
-varimp_fig
+#varimp_fig
 ggsave(here::here("figures/varImpPlot.jpg"), varimp_fig, width = 7.5, 
        height = 5.625, units = "in", dpi = 600)
 
@@ -57,7 +57,7 @@ pp_shoreline_length <- data.frame(partplot_shoreline_leng,
                                   stringsAsFactors = FALSE)
 pp_data <- rbind(pp_date, pp_avg_temp, pp_long, pp_thirty_day, pp_elev, pp_lat,
                  pp_surf_area, pp_shoreline_length)
-write_csv(pp_data, "figures/pp_data.csv")
+write_csv(pp_data, here::here("figures/pp_data.csv"))
 
 pp_data <- pp_data %>%
   mutate(variable = factor(variable, levels = c("Date", "Avg. temperature",
@@ -68,7 +68,7 @@ pp_data <- pp_data %>%
                                                 "Lake shoreline length")))
 
 pp_fig <- partial_plot(pp_data)
-pp_fig
+#pp_fig
 ggsave(here::here("figures/partPlot.jpg"), pp_fig, width = 8, 
        height = 10, units = "in", dpi = 300)
 

### 2020-06-27 R/varsel_plot.R

@@ -6,8 +6,8 @@
 #' @export
 #' @import ggplot2
 varsel_plot <- function(vs_reg_obj){
-  dat <- data.frame(num = vs_reg_obj$num_var, mse = vs_reg_obj$mse)
-  #browser()
+  dat <- data.frame(num = unlist(vs_reg_obj$num_var), mse = unlist(vs_reg_obj$mse))
+  
   plot_out<-ggplot(dat, aes(x=num,y=mse)) +
     geom_point(size = 3) +
     theme_ipsum_rc(base_size = 12, axis_title_size = 12) +

### 2020-07-06 manuscript/manuscript.Rmd
Edit to MS

@@
 author:
-- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead, W. Bryan. ^*1*^,
-- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
-  Ecology Division, Narragansett, RI 02882
+- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
+  W. Bryan. ^*1*^,
+- ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
+  Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
 - ^*2*^ ORISE, Narragansett, RI 02882
 - '^\*^  *corresponding author: kreakie.betty@epa.gov*'
-
 bibliography: manuscript.bib
@@
     reference_docx: manuscript_template.docx
+  pdf_document: default
 editor_options:
@@
 csl: plos.csl
-
-abstract: Text for abstract
-
+abstract: As the average global air temperature on Earth increases, surface temperatures
+  of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).  The
+  influence of this increased temperature touches all biotic and abiotic components
+  of lentic ecosystems. Here we present a simple yet robust model of lake photic zone
+  temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for
+  the conterminous United States. The final model has a mean square error of 2.19
+  and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average ambient air temperature,
+  data obtained from the PRISM Climate Group, and longitude are the most important
+  variables impacting the final model’s accuracy. Given the importance of temperature
+  to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can
+  be a valuable tool for researchers and lake resource managers. Daily predicted lake
+  photic zone temperature for all lakes in the conterminous US can now be straightforwardly
+  estimated based on basic ambient temperature and location information.
 ---
@@
 
-During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].
+During a time of unprecedented environmental and climatic variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].
 As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)[@hansen2010global], surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009) [@o2015rapid].
@@
 The 2017 data are currently undergoing quality control before being released the public.
+Upon release, the data can easily be included to improve this model.
 For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).
@@
 Figure 1: Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes
+
+We included numerous predictor variables that are hypothesized to impact lake photic zone.
+As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).
+This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).
+PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).
+For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.
 
@@
 Lakes with partials buffers falling outside the US were excluded.
-
-We included numerous predictor variables that are hypothesized to impact lake photic zone.
-As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).
-This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).
-PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).
-For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.
 
@@
 Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
-Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.(ADD SENTENCE)  All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
+Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
 Model performance is reported as mean square error and adjusted R.
@@
 Shows percent increase in mean square error as a function of the number of variables.", display = "cite")`).
-The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.
+The selected variables were average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.
 
@@
 
-The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R^2 of 0.88.
+The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R(n<sup>2</sup>) of 0.88.
 The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.
@@
 Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States.
-The final model has a mean square error of 2.19 and an adjusted R^2 of 0.88.
+The final model has a mean square error of 2.19 and an adjusted R(n<sup>2</sup>) of 0.88.
 The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.

### 2020-08-07 manuscript/manuscript.Rmd
new figure

@@
 output:
+  pdf_document: default
   word_document:
     reference_docx: manuscript_template.docx
-  pdf_document: default
 editor_options:
@@
 
+![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
 Figure 1: Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes

### 2020-08-07 manuscript/manuscript.tex
new figure

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Lake Photic Zone Temperature across the Conterminous United States},
  pdfauthor={Kreakie, B. J. * 1, Shivers, S. 2, Hollister. J. W. 1, Milstead, W. Bryan. 1,; 1 US Environmental Protection Agency, Office of Research and Development, Atlantic Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882; 2 ORISE, Narragansett, RI 02882; * *corresponding author: kreakie.betty@epa.gov*},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Lake Photic Zone Temperature across the Conterminous United States}
\author{Kreakie, B. J. \textsuperscript{*} \textsuperscript{\emph{1}}, Shivers,
S. \textsuperscript{\emph{2}}, Hollister. J. W.
\textsuperscript{\emph{1}}, Milstead, W. Bryan.
\textsuperscript{\emph{1}}, \and \textsuperscript{\emph{1}} US Environmental Protection Agency, Office of
Research and Development, Atlantic Coastal Environmental Sciences
Division (ACESD), Narragansett, RI 02882 \and \textsuperscript{\emph{2}} ORISE, Narragansett, RI 02882 \and \textsuperscript{*} *corresponding author:
\href{mailto:kreakie.betty@epa.gov*}{\nolinkurl{kreakie.betty@epa.gov*}}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
As the average global air temperature on Earth increases, surface
temperatures of lakes are also increasing globally (0.34 °C per decade
from 1985 to 2009). The influence of this increased temperature touches
all biotic and abiotic components of lentic ecosystems. Here we present
a simple yet robust model of lake photic zone temperature using the 2007
and 2012 US EPA's National Lakes Assessment data for the conterminous
United States. The final model has a mean square error of 2.19 and an
adjusted R(n2) of 0.88. The sampling date, that day's average ambient
air temperature, data obtained from the PRISM Climate Group, and
longitude are the most important variables impacting the final model's
accuracy. Given the importance of temperature to a lake ecosystem,
especially to cyanobacteria bloom dynamics, this model can be a valuable
tool for researchers and lake resource managers. Daily predicted lake
photic zone temperature for all lakes in the conterminous US can now be
straightforwardly estimated based on basic ambient temperature and
location information.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

During a time of unprecedented environmental and climatic variability,
lakes can serve as sentinels and integrators in a changing world
{[}1,2{]}. As the average global air temperature on Earth increases
(0.15-0.20 °C per decade since 1975){[}3{]}, surface temperatures of
lakes are also increasing globally (0.34 °C per decade from 1985 to
2009) {[}4{]}. The influence of this increased temperature touches all
biotic and abiotic components of lentic ecosystems. Ultimately,
temperature changes will greatly impact every aspects of lake resource
management. For example, temperature, in addition to nutrients, is a key
driver to cyanobacteria bloom dynamics {[}5{]}. During periods of higher
temperature, cyanobacteria species dominate the phytoplankton community
{[}6--8{]}. As lake temperature increase (typically above 25°C),
cyanobacteria have a competitive advantage over phytoplankton and can
proliferate quickly {[}9{]}. Moreover, experimentally enhanced water
temperatures yielded significantly increased growth rates of higher
toxic Microcystis, but not the non-toxic strains {[}10{]}. Thus, our
ability to understand and predict toxic cyanobacteria blooms will be
deeply dependent on our ability to forecast lake photic zone
temperature. This need for accurate lake temperature forecasting will be
crucial for protecting human and environmental health.

Because to the ecological significance, it is not surprising that
modelling near-surface lake temperature has been broadly investigated.
Models typically vary in number of lakes studied, complexity of
modelling approach, and study interval. Modelling efforts include
efforts to model a single lake predicted over relatively small-time
intervals (like hourly predictions). Such studies include: {[}11--13{]}.
There are numerous studies that model temperature for a small number of
lakes while attempting to limit the number of predictor variables
{[}14--17{]}. By and large, air temperature and lake size are often the
only selected predictor variables. On the whole, very few modelling
efforts attempt to predict lake temperature across large spatial extents
for large number of lakes. Yet some examples include: {[}4,19,20{]}.
Despite this great variation in approach and execution, most have been
rather successful modelling lake photic zone temperature, although it is
important to note that the majority of these lakes are large.
Considering that the number and importance of small lakes (\textless{}
1km2) has been historically underestimated, it is important to also
predict temperatures in these small lakes {[}21,22{]}.

In summary, we still lack the ability to robustly model near-surface
temperature for all lakes across a large spatial extent. Modelling lake
temperature requires a large amount of data, and therefore study lakes
are often selected opportunistically, which may introduce a spatial
bias. Frequently, the study lakes have a high regional resource value
(e.g.~the Great Lakes) and commonly have an extensive monitoring history
due to the vested interest of the public. When model efforts do attempt
to cross large spatial extents, these efforts often rely on satellite
data. While these models predict over large areas, they are restricted
by the size of the lake captured by satellite (typically 3km2 for 1 km
MODIS pixels).

The presented modelling effort uses the US Environmental Protection
Agency's (EPA's) National Lake Assessment (NLA). The NLA is a stratified
random sample of all lakes in the conterminous United States repeated
every five years beginning in 2007. Even though this is a large effort
involving numerous agencies, the sampling methods are standardized and
have a comprehensive quality assurance plan. The uniqueness of this data
set allows us to build a robust lake photic zone temperature model for
all US lakes. By using lakes across the US (i.e.~at a large spatial
extent), we included lakes with different morphologies, in different
climates with diverse geologies, and surrounding landscapes.
Additionally, small lakes were well represented with more than 50\% of
the lakes in the 2012 survey less than 0.5 km2.

To summarize, the main goal of this work was to develop a simple yet
robust lake photic temperature model for all lakes in the conterminous
US that capture key drivers of near-surface lake temperature.
Additionally, this work (something about open science and GitHub link).

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\#\#Data We relied on the in situ temperature data provided in the US
EPA's National Lake Assessment (NLA) (CITE). The NLA is a stratified
random sample of lakes (great than 1 ha) across the United States. NLA
sampling took place in 2007, 2012, and 2017. This research effort used
the 2007 and 2012 sample years. The 2017 data are currently undergoing
quality control before being released the public. Upon release, the data
can easily be included to improve this model. For both sample years, we
have included over 1,000 lakes across the conterminous US excluding the
Great Lakes (Figure 1). For each sampled lake, we used the mean
temperature for all sampled depths of less than 2m (this being the depth
typically considered the photic zone).

\includegraphics{../figures/LakePhoticZonev2.jpg} Figure 1: Map of 2007
and 2012 US EPA's National Lake Assessment (NLA) Lakes

We included numerous predictor variables that are hypothesized to impact
lake photic zone. As a proxy for directly measured ambient air
temperature, we used the PRISM AN81d dataset (PRISM Climate Group,
Oregon State University, \url{http://prism.oregonstate.edu}, created 07
Nov 2018). This dataset provides interpolated daily temperature
estimates (mean, maximum, and minimum) for 4km grids in the conterminous
United States for 1981 to the present (see:
\url{http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf}).
PRISM takes advantage of measured climate variables to interpolate point
data to spatially defined grids using regression techniques and expert
knowledge (Daly et al 2008). For our study, we used the ``prism'' R
package (Hart \& Bell) to download the mean daily temperatures for the
PRISM grid cells corresponding to the centroids of all NLA lakes
included in this study for dates between 01 May and 30 September of 2007
and 2012.

In addition to climate, there are additional factors clearly impacting
lake temperature (e.g., surrounding land use, lake depth, size, and
configuration, and elevation). To test for the relative importance of
lake morphometry and surrounding landscape, we used the R packages
\texttt{lakemorpho} to calculate a suite of lake morphometry metric and
\texttt{elevatr} to access digital elevation models for each lake
{[}23,24{]}. Additionally, National Land Cover Database (NLCD) was our
source for landcover data. Specifically, we calculated the percent
impervious surface of a 3000m buffer for each lake. Lakes with partials
buffers falling outside the US were excluded.

To test the relative influence of both short and long-term temperature,
we derived several measures for a lake's local air temperature. Mean air
temperatures for day of and the day before the sample date were
extracted directly from the prism data. To understand longer term
influences we calculated average mean air temperatures for periods 3, 7,
and 30 days prior to the sample date.

\hypertarget{random-forest-modelling}{%
\subsection{Random Forest Modelling}\label{random-forest-modelling}}

Random forest modelling was used to not only develop a predictive model
of photic zone temperature, but also used as a means of variable
selection and calculate relative variable importance. Random forest is a
machine learning method that builds a consensus prediction from the
assemblage of multiple tree models (here specifically 10,000 trees for
the final model and 1,000 trees for the variable selection models). Each
individual tree model is constructed from a subset of the full data set
and a subset of all predictor variables.All random forest modelling was
conducted in R v XXXX (CITE) with the randomForest package (CITE). Model
performance is reported as mean square error and adjusted R.

Random forest does not require that users reduce the number of predictor
variables; the random forest algorithm prevents overfitting and is not
impacted by correlated predictor variables (Culter 2007). While it is
unlikely that random forest models constructed with reduced numbers of
predictor variables perform any better than models constructed with a
full suite of available variables(Fox et al., 2017), reducing the number
of our predictor variables was necessary in order to fulfill the
long-term goals of this work. Several of our climatic predictor
variables are computationally intensive to create for the entire
conterminous United States. In order to use our final model for future
forecasting or historical backcasting, we strove to create a robust
predictive model while minimizing computational demands. To determine
the optimal number and set of variables, we followed the variable
selection method presented in (Hollister 2016).

In addition to measure of overall model performance, we used percent
increase of mean-squared error to assess variable importance. The
percent increase in mean-squared error is a comparison between the true
values of a variable and randomly permuted values of a specific variable
on overall model performance.

\hypertarget{results}{%
\section{Results}\label{results}}

We used 1185 data points from the 2007 NLA and 1097 from 2012 NLA.
Figure 1 illustrates the spatial distribution of the sampled lakes.

Figure 1

Using the average temperature for the upper two meters as the response
variable, we initially began the variable selection process with 16
predictor variables. The variable selection process identified a reduced
model with 7 variables (Figure 2). The selected variables were average
ambient air temperature for the sample date, sample date, longitude,
average ambient air temperature for 30 day proceeding the sample date,
elevation, latitude, length of lake shoreline, and the lake surface
area.

\begin{figure}
\centering
\includegraphics{../figures/varselfig.jpg}
\caption{Figure 2: Variable selection plot for all variables. Shows
percent increase in mean square error as a function of the number of
variables.}
\end{figure}

The final model built with the eight selected variables has a
mean-squared error of 2.19 and adjusted R(n2) of 0.88. The variables
ranked in order of importance were date, average temperature, longitude,
30-day average temperature, elevation, latitude, surface area, and
shoreline length (Figure 3). The partial dependency plots illustrate how
the predicted photic zone temperature changes over the range of values
for all predictor variables (Figure 4).

\begin{figure}
\centering
\includegraphics{../figures/varImpPlot.jpg}
\caption{Figure 3: Variable importance plot for selected variables.
Shows precent increase in mean square error. Higher values indicate a
higher impact on overall model accuracy.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/partPlot.jpg}
\caption{Figure 4: Partial dependence plots for selected variables.}
\end{figure}

\hypertarget{discussion-and-conclusions}{%
\section{Discussion and conclusions}\label{discussion-and-conclusions}}

Here we present a simple yet robust model of lake photic zone
temperature using the 2007 and 2012 NLA data for the conterminous United
States. The final model has a mean square error of 2.19 and an adjusted
R(n2) of 0.88. The sampling date, that day's average ambient air
temperature, data obtained from the PRISM Climate Group, and longitude
are the most important variables impacting the final model's accuracy.
Given the importance of temperature to a lake ecosystem, especially to
cyanobacteria bloom dynamics, this model can be a valuable tool for
researchers and lake resource managers. Daily predicted lake photic zone
temperature for all lakes in the conterminous US can now be
straightforwardly estimated based on basic ambient temperature and
location information.

Using only easily obtainable metrics to estimate lake temperature was a
primary focus in the creation of this model. Several studies
(e.g.~{[}25{]}) have used similar metrics with good results (RMSE=0.5
°C) to estimate lake temperature but were applied to a single, or a
small number of lakes. Other studies have used more complex models
(e.g.~GLM) to yield good results (RMSE = 1.62 °C for the epilimnion) at
varying scales {[}26,27{]}. However, these models require more complex
input variables which are not available for all lakes and may be biased
towards large lakes. Using the NLA data and simple metrics allowed this
model to be applicable to any lake within the conterminous United States
using minimal input data with a small increase in RMSE. Additionally,
this model is predictive and the model and data will be made publicly
available.

The final model included, average ambient air temperature and the
average temperature of the prior 30 days. Yet the 3-day and 7-day prior
to the sampling date averages were not selected in the final model. It
is likely that the three and seven day averages are not providing the
model with unique information. Typical the three and seven day averages
are fairly close the day-of average. Ambient air temperature does not
change drastically over such a short time scale. Even if a big swing in
temperature did happen during this short time period, it would be too
rare to be a significant factor in the model. Yet the longer term 30-day
average does have an impact on lake photic zone temperature. The 30-day
average provides us with this information about the temperature
intensity leading up to the sample date. Not just how far along into
summer; that information would be captured in the date parameter. But
more specifically, this measures the long term thermal heating happening
at a site.

In addition to the several derived air temperature variables, we
included land-use/land cover variables in our initial variable selection
process. Specifically, we calculated the percent impervious surface for
a 3km lake buffer and a measure of shoreline development. These
variables were included based on the hypothesis that higher amounts of
development and therefore impervious surface surrounding a lake would
lead to higher temperatures in lakes. Yet neither of these variables
were selected in the final model. Even though the land-use variables
were not selected that does not mean development and impervious are not
impactful. This urban-heat effect on lakes may have been adequately
captured in the average ambient air temperature. Therefore, making the
land-use variables redundant. Regardless these variables did not
independently contribute to the model's accuracy.

Despite being one of the most common measurements collected by
limnologists, lake temperature datasets that cover long periods of time
are very difficult to obtain. Sharma et al (2015) have compiled summer
lake temperature data for 291 lakes for the period 1985-2009. This may
be the largest lake temperature database to date, however, the data are
only available to members of their research group and, realistically,
the number of lakes included is very small. One of the reasons we chose
to model lake photic zone temperature was to develop a database of lake
temperatures for the 48 conterminous United States. The model we present
has proven to be accurate and will allow us to backcast lake
temperatures for all the \textgreater{} 300,000 lakes included in
NHDplus for the period of time covered by the PRISM climate predictions
(1981 to present). This dataset will allow us to investigate how photic
zone temperatures vary both spatially and temporally across the United
States. This database is being developed and, when complete, will be
made available as an open source data set.

\hypertarget{bibliography}{%
\section*{Bibliography}\label{bibliography}}
\addcontentsline{toc}{section}{Bibliography}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-williamson2009lakes}{}%
1. Williamson CE, Saros JE, Vincent WF, Smol JP (2009) Lakes and
reservoirs as sentinels, integrators, and regulators of climate change.
Limnology and Oceanography 54: 2273--2282.

\leavevmode\hypertarget{ref-schindler2009lakes}{}%
2. Schindler D (2009) Lakes as sentinels and integrators for the effects
of climate change on watersheds, airsheds, and landscapes. Limnology and
Oceanography 54: 2349--2358.

\leavevmode\hypertarget{ref-hansen2010global}{}%
3. Hansen J, Ruedy R, Sato M, Lo K (2010) Global surface temperature
change. Reviews of Geophysics 48.

\leavevmode\hypertarget{ref-o2015rapid}{}%
4. O'Reilly CM, Sharma S, Gray DK, Hampton SE, Read JS, et al. (2015)
Rapid and highly variable warming of lake surface waters around the
globe. Geophysical Research Letters 42: 10--773.

\leavevmode\hypertarget{ref-paerl2012climate}{}%
5. Paerl HW, Paul VJ (2012) Climate change: Links to global expansion of
harmful cyanobacteria. Water research 46: 1349--1363.

\leavevmode\hypertarget{ref-lurling2013comparison}{}%
6. Lürling M, Eshetu F, Faassen EJ, Kosten S, Huszar VL (2013)
Comparison of cyanobacterial and green algal growth rates at different
temperatures. Freshwater Biology 58: 552--559.

\leavevmode\hypertarget{ref-o2012rise}{}%
7. O'neil J, Davis T, Burford M, Gobler C (2012) The rise of harmful
cyanobacteria blooms: The potential roles of eutrophication and climate
change. Harmful algae 14: 313--334.

\leavevmode\hypertarget{ref-peperzak2003climate}{}%
8. Peperzak L (2003) Climate change and harmful algal blooms in the
north sea. Acta Oecologica 24: S139--S144.

\leavevmode\hypertarget{ref-paerl2008blooms}{}%
9. Paerl HW, Huisman J (2008) Blooms like it hot. Science 320: 57--58.

\leavevmode\hypertarget{ref-davis2009effects}{}%
10. Davis TW, Berry DL, Boyer GL, Gobler CJ (2009) The effects of
temperature and nutrients on the growth and dynamics of toxic and
non-toxic strains of microcystis during cyanobacteria blooms. Harmful
algae 8: 715--725.

\leavevmode\hypertarget{ref-saeed2016water}{}%
11. Saeed S, Honeyeh K, Ozgur K, Wen-Cheng L (2016) Water temperature
prediction in a subtropical subalpine lake using soft computing
techniques. Earth Sciences Research Journal 20: 1--11.

\leavevmode\hypertarget{ref-peeters2002modeling}{}%
12. Peeters F, Livingstone DM, Goudsmit G-H, Kipfer R, Forster R (2002)
Modeling 50 years of historical temperature profiles in a large central
european lake. Limnology and Oceanography 47: 186--197.

\leavevmode\hypertarget{ref-zhong2016recent}{}%
13. Zhong Y, Notaro M, Vavrus SJ, Foster MJ (2016) Recent accelerated
warming of the laurentian great lakes: Physical drivers. Limnology and
Oceanography 61: 1762--1786.

\leavevmode\hypertarget{ref-matuszek1996empirical}{}%
14. Matuszek JE, Shuter BJ (1996) An empirical method for the prediction
of daily water temperatures in the littoral zone of temperate lakes.
Transactions of the American Fisheries Society 125: 622--627.

\leavevmode\hypertarget{ref-kettle2004empirical}{}%
15. Kettle H, Thompson R, Anderson NJ, Livingstone DM (2004) Empirical
modeling of summer lake surface temperatures in southwest greenland.
Limnology and Oceanography 49: 271--282.

\leavevmode\hypertarget{ref-piccolroaz2016prediction}{}%
16. Piccolroaz S (2016) Prediction of lake surface temperature using the
air2water model: Guidelines, challenges, and future perspectives.
Advances in Oceanography and Limnology.

\leavevmode\hypertarget{ref-toffolon2014prediction}{}%
17. Toffolon M, Piccolroaz S, Majone B, Soja A-M, Peeters F, et al.
(2014) Prediction of surface temperature in lakes with different
morphology using air temperature. Limnology and Oceanography 59:
2185--2202.

\leavevmode\hypertarget{ref-livingstone1998relationship}{}%
18. Livingstone DM, Lotter AF (1998) The relationship between air and
water temperatures in lakes of the swiss plateau: A case study with
pal\(\backslash\)sgmaelig; olimnological implications. Journal of
Paleolimnology 19: 181--198.

\leavevmode\hypertarget{ref-minns2017factors}{}%
19. Minns CK, Shuter BJ, Davidson A, Wang S (2017) Factors influencing
peak summer surface water temperature in canada's large lakes. Canadian
Journal of Fisheries and Aquatic Sciences 75: 1005--1018.

\leavevmode\hypertarget{ref-wan2017comprehensive}{}%
20. Wan W, Li H, Xie H, Hong Y, Long D, et al. (2017) A comprehensive
data set of lake surface water temperature over the tibetan plateau
derived from modis lst products 2001--2015. Scientific data 4: 170095.

\leavevmode\hypertarget{ref-downing2006global}{}%
21. Downing JA, Prairie Y, Cole J, Duarte C, Tranvik L, et al. (2006)
The global abundance and size distribution of lakes, ponds, and
impoundments. Limnology and Oceanography 51: 2388--2397.

\leavevmode\hypertarget{ref-winslow2014lake}{}%
22. Winslow LA, Read JS, Hanson PC, Stanley EH (2014) Lake shoreline in
the contiguous united states: Quantity, distribution and sensitivity to
observation resolution. Freshwater biology 59: 213--223.

\leavevmode\hypertarget{ref-hollister2017lakemorpho}{}%
23. Hollister J, Stachelek J (2017) Lakemorpho: Calculating lake
morphometry metrics in r. F1000Research 6.

\leavevmode\hypertarget{ref-hollister2017elevatr}{}%
24. Hollister J, Tarak Shah (2017) Elevatr: Access elevation data from
various apis. Available: \url{http://github.com/usepa/elevatr}.

\leavevmode\hypertarget{ref-piccolroaz2018predictability}{}%
25. Piccolroaz S, Healey N, Lenters J, Schladow S, Hook S, et al. (2018)
On the predictability of lake surface temperature using air temperature
in a changing climate: A case study for lake tahoe (usa). Limnology and
Oceanography 63: 243--261.

\leavevmode\hypertarget{ref-bruce2018multi}{}%
26. Bruce LC, Frassl MA, Arhonditsis GB, Gal G, Hamilton DP, et al.
(2018) A multi-lake comparative analysis of the general lake model
(glm): Stress-testing across a global observatory network. Environmental
Modelling \& Software 102: 274--291.

\leavevmode\hypertarget{ref-hipsey2019general}{}%
27. Hipsey MR, Bruce LC, Boon C, Busch B, Carey CC, et al. (2019) A
general lake model (glm 3.0) for linking with high-frequency sensor data
from the global lake ecological observatory network (gleon).

\end{document}

### 2020-08-07 R/model.R

@@ -36,6 +36,13 @@ set.seed(42)
 RFAll <- randomForest(tmean_2m ~ ., data = nla_select, importance = TRUE,
                ntree = 10000)
 
+
+###Predict RF values for all points
+RFAll.predict<- predict(RFAll, data=nla_select[,2:9])
+RFAll.combined<-cbind(nla_select,RFAll.predict)
+write.csv(file="RFAlllPredict.csv",RFAll.combined)
+reg <- lm(RFAll.combined[,10]~RFAll.combined[,1], data = RFAll.combined)
+
 ## Calculate partial dependencies
 
 

### 2020-08-26 manuscript/manuscript.Rmd
edits made to Rmd and back to Betty

@@
 output:
-  pdf_document: default
   word_document:
     reference_docx: manuscript_template.docx
+  pdf_document: default
+  html_document:
+    df_print: paged
 editor_options:
@@
   the conterminous United States. The final model has a mean square error of 2.19
-  and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average ambient air temperature,
-  data obtained from the PRISM Climate Group, and longitude are the most important
-  variables impacting the final model’s accuracy. Given the importance of temperature
-  to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can
-  be a valuable tool for researchers and lake resource managers. Daily predicted lake
-  photic zone temperature for all lakes in the conterminous US can now be straightforwardly
-  estimated based on basic ambient temperature and location information.
+  and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average
+  ambient air temperature, data obtained from the PRISM Climate Group, and longitude
+  are the most important variables impacting the final model’s accuracy. Given the
+  importance of temperature to a lake ecosystem, especially to cyanobacteria bloom
+  dynamics, this model can be a valuable tool for researchers and lake resource managers.
+  Daily predicted lake photic zone temperature for all lakes in the conterminous US
+  can now be straightforwardly estimated based on basic ambient temperature and location
+  information.
 ---
@@
 During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].
-As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].
+As lake temperature increases (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].
 Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].
@@
 
-Because to the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.
+Because of the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.
 Models typically vary in number of lakes studied, complexity of modelling approach, and study interval.
@@
 This research effort used the 2007 and 2012 sample years.
-The 2017 data are currently undergoing quality control before being released the public.
+The 2017 data are currently undergoing quality control before being released to the public.
 Upon release, the data can easily be included to improve this model.
@@
 In addition to climate, there are additional factors clearly impacting lake temperature (e.g., surrounding land use, lake depth, size, and configuration, and elevation).
-To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages `lakemorpho` to calculate a suite of lake morphometry metric and `elevatr` to access digital elevation models for each lake [@hollister2017lakemorpho; @hollister2017elevatr].
+To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages `lakemorpho` to calculate a suite of lake morphometry metrics and `elevatr` to access digital elevation models for each lake [@hollister2017lakemorpho; @hollister2017elevatr].
 Additionally, National Land Cover Database (NLCD) was our source for landcover data.
 Specifically, we calculated the percent impervious surface of a 3000m buffer for each lake.
-Lakes with partials buffers falling outside the US were excluded.
+Lakes with partial buffers falling outside the US were excluded.
 
@@
 
-Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.
+Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and to calculate relative variable importance.
 Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
@@
 
-The final model included, average ambient air temperature and the average temperature of the prior 30 days.
+The final model included average ambient air temperature and the average temperature of the prior 30 days.
 Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.
@@
 
+# Alternate to the above paragraph
+
+The final model included average ambient air temperature and the average temperature of the prior 30 days,yet the 3-day and 7-day prior to the sampling date averages were not selected.
+It is likely that the three and seven day averages are not providing the model with unique information because of the focus on photic zone temperature.
+The upper 2m of the water column is more responsive to short term temperature changes compared to the whole lake, thus the 3 and 7 day temperatures are likely redundant with the sample date temperature.
+Yet the longer term 30-day average does have an impact on lake photic zone temperature, more specifically, this measures the long term thermal heating happening at a site.
+The 30-day average provides us with information about the temperature intensity leading up to the sample date.
+This is a separate measure from how far along a lake is into summer; that information would be captured in the date parameter.
+
 In addition to the several derived air temperature variables, we included land-use/land cover variables in our initial variable selection process.
@@
 These variables were included based on the hypothesis that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.
-Yet neither of these variables were selected in the final model.
+Yet neither of these variables were selected in the final model, although lake shoreline length was selected.
 Even though the land-use variables were not selected that does not mean development and impervious are not impactful.
@@
 Regardless these variables did not independently contribute to the model’s accuracy.
+It is important to note that including other variables that impact lake temperature, such as solar radiation, would likely improve the accuracy of the model.
+However, these variables are not available for all lakes in the conterminous United States.
 

### 2020-08-28 manuscript/manuscript.Rmd
Small edits to manuscript

--- 2020-07-06 manuscript/manuscript.Rmd
@@
 
-In summary, we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
+Yet we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
 Modelling lake temperature requires a large amount of data, and therefore study lakes are often selected opportunistically, which may introduce a spatial bias.
@@
 Sharma et al (2015) have compiled summer lake temperature data for 291 lakes for the period 1985-2009.
-This may be the largest lake temperature database to date, however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.
+This may be the largest lake temperature database to date; however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.
 One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.

### 2020-09-28 manuscript/manuscript.Rmd
almost done with intro read/edits

@@
 ---
-title: "Lake Photic Zone Temperature across the Conterminous United States"
+title: "**Title:** Lake Photic Zone Temperature across the Conterminous United States"
 author:
-- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
-  W. Bryan. ^*1*^,
 - ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
@@
 - ^*2*^ ORISE, Narragansett, RI 02882
+- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
+  W. Bryan. ^*1*^'
 - '^\*^  *corresponding author: kreakie.betty@epa.gov*'
@@
   word_document:
-    reference_docx: manuscript_template.docx
+    reference_docx: manuscript_template_submission.docx
   pdf_document: default
-editor_options:
-  chunk_output_type: inline
-csl: plos.csl
-abstract: As the average global air temperature on Earth increases, surface temperatures
-  of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).  The
-  influence of this increased temperature touches all biotic and abiotic components
-  of lentic ecosystems. Here we present a simple yet robust model of lake photic zone
-  temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for
-  the conterminous United States. The final model has a mean square error of 2.19
-  and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average ambient air temperature,
-  data obtained from the PRISM Climate Group, and longitude are the most important
-  variables impacting the final model’s accuracy. Given the importance of temperature
-  to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can
-  be a valuable tool for researchers and lake resource managers. Daily predicted lake
-  photic zone temperature for all lakes in the conterminous US can now be straightforwardly
-  estimated based on basic ambient temperature and location information.
+csl: ecology.csl
+abstract: |
+  | **Abstract:** As the average global air temperature on Earth increases, surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).
+The influence of this increased temperature is known to impact lentic ecosystems across local to broad scales.
+For example, warming lake temperature is linked to disruptions in trophic linkages, changes in thermal stratification, and cyanobacteria bloom dynamics.
+Thus understanding broad trends in lake temperature is important to understanding the changing ecology of lakes and the potential human health impacts of these changes.
+To help address this, we developed a simple yet robust model of lake photic zone temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for the conterminous United States.
+The final model has a mean square error of 2.19 and an adjusted R^2^ of 0.88.
+The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
+Given the importance of temperature to a lake ecosystem, this model can be a valuable tool for researchers and lake resource managers.
+Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
+  | **Keywords:** Random forest, National Lakes Assessment, Limnology, water temperature, warming lakes
 ---
@@
 The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.
-Ultimately, temperature changes will greatly impact every  aspects of lake resource management.
-For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].
+
+For example, warming lakes are linked to a disruption in trophic linkages between phytoplankton and zooplankton [@winder2004climate].
+Specifically, Winder and Schindler [-@winder2004climate] report a spring diatom bloom occurring 20 days earlier and a long-term decline in Daphnia populations in Lake Washington, a lake near Seattle, Washington.
+Second, warming has been shown to result in changes in thermal stratification across different lake types [@hampton2018change; @oreilly2003climate; @verburg2003ecological; @cohen2016climate; michelutti2016equatorial].
+In particular, lakes that typically stratify can exhibit a strengthening of the stratification as well as a reduction in the depth of the thermocline or previously polymictic lakes can begin to stratify as a result of warming lake temperatures.
+Temperature, in addition to nutrients, is also a key driver to cyanobacteria bloom dynamics [@paerl2012climate].
 During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].
@@
 Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.
-This need for accurate lake temperature forecasting will be crucial for protecting human and environmental health.
+Ultimately, temperature changes will greatly impact every aspect of lake ecology and lake resource management.
 
-Because to the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.
+Because of the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.
 Models typically vary in number of lakes studied, complexity of modelling approach, and study interval.
-Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).
-Such studies include: [@saeed2016water; @peeters2002modeling; @zhong2016recent].
-There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
-By and large, air temperature and lake size are often the only selected predictor variables.
-On the whole, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.
-Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].
-Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.
-Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes [@downing2006global; @winslow2014lake].
+Modelling efforts include efforts to model a single lake over relatively small (e.g.
+hourly) time intervals [@saeed2016water; @peeters2002modeling; @zhong2016recent].
+Additionally, there are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
+In these efforts, air temperature and lake size are often the only selected predictor variables.
+There have also been some efforts aimed at modelling and measuring lake temperature across large spatial extents for large number of lakes [@minns2017factors; @o2015rapid; @wan2017comprehensive].
+These past efforts use a wide variety of approaches and have been rather successful at modelling lake photic zone temperature.
+One limitation, however, is that the majority of these lakes are large and few studies document or model smaller lakes.
+Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these smaller waterbodies [@downing2006global; @winslow2014lake].
 
-Yet we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
+In spite of the need to understand near-surface temperature for all lakes across a large spatial extent, there have been several challenges that have slowed our progress.
 Modelling lake temperature requires a large amount of data, and therefore study lakes are often selected opportunistically, which may introduce a spatial bias.
@@
 
-The presented modelling effort uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
+Our modelling effort takes advantage of the relatively recent availability of broad scale field data for lakes and uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
 The NLA is a stratified random sample of all lakes in the conterminous United States repeated every five years beginning in 2007.
@@
 
-To summarize, the main goal of this work was to develop a simple yet robust lake photic temperature model for all lakes in the conterminous US that capture key drivers of near-surface lake temperature.
+The main goals of this work are to 1) develop a simple yet robust lake photic temperature model, 2) develop a model that is applicable to all lakes in the conterminous US that capture key drivers of near-surface lake temperature, and 3.
 Additionally, this work (something about open science and GitHub link).

### 2020-09-28 manuscript/manuscript.Rmd
added some cites, fixed fig 1 captions

@@
 
-The main goals of this work are to 1) develop a simple yet robust lake photic temperature model, 2) develop a model that is applicable to all lakes in the conterminous US that capture key drivers of near-surface lake temperature, and 3.
-Additionally, this work (something about open science and GitHub link).
+The main goals of this work are to 1) develop a simple yet robust lake photic zone temperature model and 2) develop a model that is applicable to all lakes in the conterminous US that capture key drivers of near-surface lake temperature.
+Additionally, it is our practice conduct our work as openly as is feasible.
+Towards that end we provide access to all code and data used to develop these models with the active repository available at <https://github.com/usepa/lake_photic_zone> and the code, data, and documents are archived at <https://link_to_zenodo_right_before_submission>.
 
@@
 
-##Data
-We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment (NLA) (CITE).
+## Data
+
+We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment [@usepa2016national; @usepa2009national].
 The NLA is a stratified random sample of lakes (great than 1 ha) across the United States.
@@
 Upon release, the data can easily be included to improve this model.
-For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).
+For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (`r fig_num("LakePhoticZonev2", caption = "Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes", display = "cite")`).
 For each sampled lake, we used the mean temperature for all sampled depths of less than 2m (this being the depth typically considered the photic zone).
 
-![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
-Figure 1: Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes
+![](../figures/LakePhoticZonev2.jpg)
+
+`r fig_num("LakePhoticZonev2")`
 

### 2020-09-29 manuscript/manuscript.Rmd
more edits

@@
 As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).
-This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).
-PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).
-For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.
+This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: <http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf>).
+PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge [@daly2008physiographically].
+For our study, we used the “prism” R package to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012 [@hart2015prism].
 
@@
 To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages `lakemorpho` to calculate a suite of lake morphometry metric and `elevatr` to access digital elevation models for each lake [@hollister2017lakemorpho; @hollister2017elevatr].
-Additionally, National Land Cover Database (NLCD) was our source for landcover data.
+Additionally, the National Land Cover Database (NLCD) was our source for land cover data.
 Specifically, we calculated the percent impervious surface of a 3000m buffer for each lake.
@@
 
-Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.
+Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used to aid in variable selection and to calculate relative variable importance.
 Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
-Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
-Model performance is reported as mean square error and adjusted R.
+Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.
+All random forest modelling was conducted in R, version 4.0.2 with the randomForest package [@R2020; @liaw2002rf].
+Model performance is reported as mean square error and adjusted R^2^.
 
-Random forest does not require that users reduce the number of predictor variables; the random forest algorithm prevents overfitting and is not impacted by correlated predictor variables (Culter 2007).
-While it is unlikely that random forest models constructed with reduced numbers of predictor variables perform any better than models constructed with a full suite of available variables(Fox et al., 2017), reducing the number of our predictor variables was necessary in order to fulfill the long-term goals of this work.
+Random forest does not require that users reduce the number of predictor variables because the random forest algorithm prevents overfitting and is not impacted by correlated predictor variables [@cutler2007random].
+It is unlikely that random forest models constructed with reduced numbers of predictor variables perform any better than models constructed with a full suite of available variables [@fox2017assessing].
+However, reducing the number of predictor variables eases interpretation and can reduce potentially unneeded computation time while also not severely impacting prediction accuracy.
 Several of our climatic predictor variables are computationally intensive to create for the entire conterminous United States.
-In order to use our final model for future forecasting or historical backcasting, we strove to create a robust predictive model while minimizing computational demands.
-To determine the optimal number and set of variables, we followed the variable selection method presented in (Hollister 2016).
+So, in order to use our final model for future forecasting or historical backcasting, we strove to create a robust predictive model while minimizing computational demands.
+To determine the optimal number and set of variables, we followed the variable selection method presented in [@hollister2016modeling].
 
-In addition to measure of overall model performance, we used percent increase of mean-squared error to assess variable importance.
+In addition to a measure of overall model performance, we used percent increase of mean-squared error to assess variable importance.
 The percent increase in mean-squared error is a comparison between the true values of a variable and randomly permuted values of a specific variable on overall model performance.
@@
 
-We used 1185 data points from the 2007 NLA and 1097 from 2012 NLA.  Figure 1 illustrates the spatial distribution of the sampled lakes.
+```{r summary, echo=FALSE}
+summ <- nla.all %>%
+  group_by(year) %>%
+  summarize(n = n())
+```
 
-Figure 1
+We used `r summ[summ$year == "2007",]$n` data points from the 2007 NLA and `r summ[summ$year == "2012",]$n` from 2012 NLA and these points are a spatially representative sample of lakes in the United States (`r fig_num("LakePhoticZonev2", display = "cite")`)
+
+![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
 
 Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.
-The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables.
+The variable selection process identified a reduced model with ~8 variables showing minimal model error (`r fig_num("varsel_fig",caption="Variable selection plot for all variables.
 Shows percent increase in mean square error as a function of the number of variables.", display = "cite")`).
@@
 
-![`r fig_num("varsel_fig")`](../figures/varselfig.jpg)
+![`r fig_num("varsel_fig")`](../figures/varSelFig.jpg)
 
-The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R(n<sup>2</sup>) of 0.88.
+The final model built with the eight selected variables has a mean-squared error of `r round(mean(RFAll$mse), 2)`  and adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
 The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.
@@
 Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States.
-The final model has a mean square error of 2.19 and an adjusted R(n<sup>2</sup>) of 0.88.
+The final model has a mean square error of `r round(mean(RFAll$mse), 2)` and an adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
 The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
 Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers.
-Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.
+Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
 
@@
 Other studies have used more complex models (e.g.
-GLM) to yield good results (RMSE = 1.62 °C for the epilimnion) at varying scales [@bruce2018multi; @hipsey2019general].
+General Lake Model (GLM)) to yield good results (RMSE = 1.62 °C for the epilimnion) at varying scales [@bruce2018multi; @hipsey2019general].
 However, these models require more complex input variables which are not available for all lakes and may be biased towards large lakes.
@@
 Despite being one of the most common measurements collected by limnologists, lake temperature datasets that cover long periods of time are very difficult to obtain.
-Sharma et al (2015) have compiled summer lake temperature data for 291 lakes for the period 1985-2009.
+Sharma et al [-@sharma2015global] have compiled summer lake temperature data for 291 lakes for the period 1985-2009.
 This may be the largest lake temperature database to date; however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.

### 2020-10-02 manuscript/manuscript.Rmd
this round of edits done...

@@
 To help address this, we developed a simple yet robust model of lake photic zone temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for the conterminous United States.
-The final model has a mean square error of 2.19 and an adjusted R^2^ of 0.88.
+The final model has a root mean square error of 1.48 °C and an adjusted R^2^ of 0.88.
 The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
@@
 library(captioner)
+library(dplyr)
 #opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
@@
 
-```{r model, include=FALSE, eval=TRUE, cache=TRUE}
+```{r model, include=FALSE, eval=TRUE}
 source(here::here("R/model.R"), encoding = "UTF-8")
 ```
-```{r figures, include=FALSE, cache=TRUE}
+```{r figures, include=FALSE}
 source(here::here("R/figures.R"), encoding = "UTF-8")
@@
 Specifically, Winder and Schindler [-@winder2004climate] report a spring diatom bloom occurring 20 days earlier and a long-term decline in Daphnia populations in Lake Washington, a lake near Seattle, Washington.
-Second, warming has been shown to result in changes in thermal stratification across different lake types [@hampton2018change; @oreilly2003climate; @verburg2003ecological; @cohen2016climate; michelutti2016equatorial].
+Second, warming has been shown to result in changes in thermal stratification across different lake types [@hampton2018change; @oreilly2003climate; @verburg2003ecological; @cohen2016climate; @michelutti2016equatorial].
 In particular, lakes that typically stratify can exhibit a strengthening of the stratification as well as a reduction in the depth of the thermocline or previously polymictic lakes can begin to stratify as a result of warming lake temperatures.
@@
 hourly) time intervals [@saeed2016water; @peeters2002modeling; @zhong2016recent].
-Additionally, there are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
+Additionally, there are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical; @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
 In these efforts, air temperature and lake size are often the only selected predictor variables.
@@
 
-![](../figures/LakePhoticZonev2.jpg)
-
-`r fig_num("LakePhoticZonev2")`
+![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
 
@@
 
-```{r summary, echo=FALSE}
+```{r summary, include=FALSE, message=FALSE}
 summ <- nla.all %>%
@@
   summarize(n = n())
+
+extreme <- nla_select %>%
+  filter(tmean_2m < 15 | tmean_2m > 32) %>%
+  summarize(perc = nrow(.)/nrow(nla_select) *100)
 ```
@@
 We used `r summ[summ$year == "2007",]$n` data points from the 2007 NLA and `r summ[summ$year == "2012",]$n` from 2012 NLA and these points are a spatially representative sample of lakes in the United States (`r fig_num("LakePhoticZonev2", display = "cite")`)
-
-![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
 
@@
 
+<!-- Check y-axis label -->
 ![`r fig_num("varsel_fig")`](../figures/varSelFig.jpg)
 
-The final model built with the eight selected variables has a mean-squared error of `r round(mean(RFAll$mse), 2)`  and adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
+The final model built with the eight selected variables has a root mean-squared error of `r round(sqrt(mean(RFAll$mse)), 2)`  and adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
+The model performs well across a wide range of temperatures (`r fig_num("obs_v_pred", caption = "Predicted versus measured photic zone temperature", display = "cite")`); however, at the higher and lower temperature (i.e.
+less than 15 °C and greater than 32 °C) the model does not perform as well but lakes with growing season temperatures at these extremes were rare and represented only `r round(extreme$perc, 2)` of lakes sampled in the 2007 and 2012 NLA.
+
+![`r fig_num("obs_v_pred")`](../figures/obs_v_pred.jpg)
+
 The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.
@@
 Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States.
-The final model has a mean square error of `r round(mean(RFAll$mse), 2)` and an adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
+The final model has a root mean square error of `r round(sqrt(mean(RFAll$mse)), 2)` and an adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
 The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
@@
 Several studies (e.g.
-[@piccolroaz2018predictability]) have used similar metrics with good results (RMSE=0.5 °C) to estimate lake temperature but were applied to a single, or a small number of lakes.
+[@piccolroaz2018predictability]) have used similar metrics with good results (RMSE = 0.5 °C) to estimate lake temperature but were applied to a single, or a small number of lakes.
 Other studies have used more complex models (e.g.
@@
 However, these models require more complex input variables which are not available for all lakes and may be biased towards large lakes.
-Using the NLA data and simple metrics allowed this model to be applicable to any lake within the conterminous United States using minimal input data with a small increase in RMSE.
-Additionally, this model is predictive and the model and data will be made publicly available.
+Using the NLA data and simple metrics allowed this model to be applicable to any lake within the conterminous United States using minimal input data with a comparable RMSE.
 
@@
 Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.
-It is likely that the three and seven day averages are not providing the model with unique information.
-Typical the three and seven day averages are fairly close the day-of average.
-Ambient air temperature does not change drastically over such a short time scale.
-Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.
-Yet the longer term 30-day average does have an impact on lake photic zone temperature.
-The 30-day average provides us with this information about the temperature intensity leading up to the sample date.
+It is likely that the three and seven day averages are not providing the model with unique information as the three and seven day averages are likely to be similar to the day-of average.
+Ambient air temperature does not change drastically over such a short time scale [CITATION??].
+Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.<!--JWH: We can quantify this with the data that we have.
+Should probably do this.
+Less arm wavey --> Yet the 30-day average does have an impact on lake photic zone temperature as it provides information about the longer term temperature intensity leading up to the sample date.
 Not just how far along into summer; that information would be captured in the date parameter.
@@
 This may be the largest lake temperature database to date; however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.
-One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.
+<!--These data are open and published - accessible through the "laketemps" package.-->  One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.
 The model we present has proven to be accurate and will allow us to backcast lake temperatures for all the > 300,000 lakes included in NHDplus for the period of time covered by the PRISM climate predictions (1981 to present).

### 2020-10-02 R/figures.R

@@ -9,6 +9,7 @@ source(here::here("R/partial_plot.R"))
 library(ggplot2)
 library(hrbrthemes)
 library(readr)
+nla_select <- read_csv(here::here("data/nla_select.csv"))
 
 ###Variable selection figure 
 varsel_fig<-varsel_plot(var_sel_tmean)
@@ -72,5 +73,20 @@ pp_fig <- partial_plot(pp_data)
 ggsave(here::here("figures/partPlot.jpg"), pp_fig, width = 8, 
        height = 10, units = "in", dpi = 300)
 
+# Obs v Pred
+obs_v_pred_gg <- data.frame(observed = nla_select$tmean_2m, 
+                            predicted = RFAll$predicted) %>%
+  ggplot(aes(x = observed, y = predicted)) +
+  geom_point(alpha = 0.35) +
+  geom_abline(slope = 1, intercept = 0, size = 1.25, color = "darkblue") +
+  theme_ipsum_rc() +
+  scale_x_continuous(limits = c(8,38), breaks = seq(10, 35, 5)) +
+  scale_y_continuous(limits = c(8,38), breaks = seq(10, 35, 5)) +
+  labs(y = "Predicted photic zone temperature (°C)", 
+       x = "Measured photic zone temperature (°C)") +
+  theme(axis.title.x = element_text(size = 12, vjust = -1),
+        axis.title.y = element_text(size = 12, vjust = 4))
 
+ggsave(here::here("figures/obs_v_pred.jpg"), obs_v_pred_gg, width = 7.5, 
+                  height = 5.625, units = "in", dpi = 600)
 

### 2020-10-02 R/model.R

@@ -5,6 +5,7 @@ library(purrr)
 library(here)
 library(broom)
 library(lubridate)
+library(readr)
 
 source(here::here('R/varsel_regression_rf.R'))
 
@@ -16,35 +17,43 @@ nla.cc$year<-factor(nla.cc$year)
 nla.cc$date <- lubridate::yday(lubridate::ymd(nla.cc$date))
 
 nla.all <-  filter(nla.cc, tmean_2m < 70) %>%
-              dplyr::select(tmean_2m, year,date, longitude, latitude, tmean,
+              dplyr::select(nla_id, tmean_2m, year,date, longitude, latitude, tmean,
               tmean_avg30,percent_impervious,surface_area,shoreline_length,
               shoreline_dev, max_length, max_width, mean_width, max_depth,
               max_depth, volume, elevation)
 
-nla.all<-rename(nla.all,Date=date,Average.Temperature=tmean,Longitude=longitude,
+nla.all<-rename(nla.all, nla_id = nla_id, Date=date,Average.Temperature=tmean,Longitude=longitude,
                  Thirty.Day.Average.Temperature=tmean_avg30,Elevation=elevation,
                  Latitude=latitude,Surface.Area=surface_area,
                  Shoreline.Length=shoreline_length)
 
 ##Variable selection for 2007&2012 model
 set.seed(42)
-var_sel_tmean <- varsel_regression_rf(nla.all$tmean_2m, select(nla.all, -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
+var_sel_tmean <- varsel_regression_rf(nla.all[,-1]$tmean_2m, select(nla.all[,-1], -tmean_2m),                ntree=1000, prog=TRUE, importance=TRUE)
 
 
-nla_select <- select(nla.all, tmean_2m,var_sel_tmean$vars[[7]])
+nla_select <- select(nla.all, nla_id, tmean_2m,var_sel_tmean$vars[[7]])
+write_csv(nla_select, here::here("data/nla_select.csv"))
 set.seed(42)
-RFAll <- randomForest(tmean_2m ~ ., data = nla_select, importance = TRUE,
+RFAll <- randomForest(tmean_2m ~ ., data = nla_select[,-1], importance = TRUE,
                ntree = 10000)
 
 
 ###Predict RF values for all points
[…]

### 2020-10-02 R/model.R
other changes

@@ -54,6 +54,8 @@ rf_all_trees <- mutate(rf_all_trees, rmse = apply(rf_all_trees, 1, function(x)
   sqrt(mean((as.numeric(x[3:length(x)]) - as.numeric(x[2]))^2))))
 rf_all_rmse <- select(rf_all_trees, nla_id, tmean_2m, rmse) %>%
   left_join(nla)
-#plot(density(rf_all_rmse$rmse))
-#peak <- rmse_dens$x[which.max(rmse_dens$y)]
-#abline(v = peak)
+rmse_dens <- density(rf_all_rmse$rmse)
+plot(rmse_dens)
+peak <- rmse_dens$x[which.max(rmse_dens$y)]
+abline(v = peak)
+abline(v = 1.5, col="Red")

### 2020-10-06 R/model.R
all tree predictions now use just the oob

@@ -36,7 +36,7 @@ nla_select <- select(nla.all, nla_id, tmean_2m,var_sel_tmean$vars[[7]])
 write_csv(nla_select, here::here("data/nla_select.csv"))
 set.seed(42)
 RFAll <- randomForest(tmean_2m ~ ., data = nla_select[,-1], importance = TRUE,
-               ntree = 10000)
+               ntree = 10000, keep.inbag = TRUE)
 
 
 ###Predict RF values for all points
@@ -45,17 +45,34 @@ write.csv(file="RFAlllPredict.csv",RFAll.combined)
 reg <- lm(RFAll.combined[,11]~RFAll.combined[,2], data = RFAll.combined)
 
 ## Predict for all RF Trees
-rf_all_trees <- predict(RFAll, newdata = nla_select[,-1], predict.all = TRUE)
-rf_all_trees <- data.frame(rf_all_trees$individual)
+get_oob_predictions <- function(rf_obj, newdata){
+  if(!"inbag" %in% names(rf_obj)){stop("The in bag matrix is not present.  Try re-running random forest with keep.inbag = T.")}
+  rf_pred <- predict(rf_obj, newdata = newdata, predict.all = TRUE)
+  rf_inbag <- rf_obj$inbag
+  rf_inbag[rf_inbag != 0] <- NA
+  rf_pred$individual + rf_inbag
+} 
+
+rf_all_trees <- get_oob_predictions(RFAll, newdata = nla_select[,-1])
+
+rf_all_trees <- data.frame(rf_all_trees)
 rf_all_trees <- mutate(rf_all_trees, nla_id = nla_select$nla_id, 
                        tmean_2m = nla_select$tmean_2m)
 rf_all_trees <- select(rf_all_trees, nla_id, tmean_2m, everything())
-rf_all_trees <- mutate(rf_all_trees, rmse = apply(rf_all_trees, 1, function(x) 
-  sqrt(mean((as.numeric(x[3:length(x)]) - as.numeric(x[2]))^2))))
+rf_all_trees <- mutate(rf_all_trees, rmse = apply(rf_all_trees[,-1], 1, function(x) 
+  sqrt(mean((x[2:length(x)] - x[1])^2, na.rm = TRUE))))
 rf_all_rmse <- select(rf_all_trees, nla_id, tmean_2m, rmse) %>%
   left_join(nla)
[…]

### 2020-10-08 manuscript/manuscript.Rmd

--- 2020-09-28 manuscript/manuscript.Rmd
@@
 ---
-title: "**Title:** Lake Photic Zone Temperature across the Conterminous United States"
+title: "Lake Photic Zone Temperature across the Conterminous United States"
 author:
+- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
+  W. Bryan. ^*1*^,
 - ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
@@
 - ^*2*^ ORISE, Narragansett, RI 02882
-- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
-  W. Bryan. ^*1*^'
 - '^\*^  *corresponding author: kreakie.betty@epa.gov*'
@@
   word_document:
-    reference_docx: manuscript_template_submission.docx
+    reference_docx: manuscript_template.docx
   pdf_document: default
-csl: ecology.csl
-abstract: |
-  | **Abstract:** As the average global air temperature on Earth increases, surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).
-The influence of this increased temperature is known to impact lentic ecosystems across local to broad scales.
-For example, warming lake temperature is linked to disruptions in trophic linkages, changes in thermal stratification, and cyanobacteria bloom dynamics.
-Thus understanding broad trends in lake temperature is important to understanding the changing ecology of lakes and the potential human health impacts of these changes.
-To help address this, we developed a simple yet robust model of lake photic zone temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for the conterminous United States.
-The final model has a mean square error of 2.19 and an adjusted R^2^ of 0.88.
-The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
-Given the importance of temperature to a lake ecosystem, this model can be a valuable tool for researchers and lake resource managers.
-Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
-  | **Keywords:** Random forest, National Lakes Assessment, Limnology, water temperature, warming lakes
+editor_options:
+  chunk_output_type: inline
+csl: plos.csl
+abstract: As the average global air temperature on Earth increases, surface temperatures
+  of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).  The
+  influence of this increased temperature touches all biotic and abiotic components
+  of lentic ecosystems. Here we present a simple yet robust model of lake photic zone
+  temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for
+  the conterminous United States. The final model has a mean square error of 2.19
+  and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average
+  ambient air temperature, data obtained from the PRISM Climate Group, and longitude
+  are the most important variables impacting the final model’s accuracy. Given the
+  importance of temperature to a lake ecosystem, especially to cyanobacteria bloom
+  dynamics, this model can be a valuable tool for researchers and lake resource managers.
+  Daily predicted lake photic zone temperature for all lakes in the conterminous US
+  can now be straightforwardly estimated based on basic ambient temperature and location
+  information.
 ---
@@
 The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.
-
-For example, warming lakes are linked to a disruption in trophic linkages between phytoplankton and zooplankton [@winder2004climate].
-Specifically, Winder and Schindler [-@winder2004climate] report a spring diatom bloom occurring 20 days earlier and a long-term decline in Daphnia populations in Lake Washington, a lake near Seattle, Washington.
-Second, warming has been shown to result in changes in thermal stratification across different lake types [@hampton2018change; @oreilly2003climate; @verburg2003ecological; @cohen2016climate; michelutti2016equatorial].
-In particular, lakes that typically stratify can exhibit a strengthening of the stratification as well as a reduction in the depth of the thermocline or previously polymictic lakes can begin to stratify as a result of warming lake temperatures.
-Temperature, in addition to nutrients, is also a key driver to cyanobacteria bloom dynamics [@paerl2012climate].
+Ultimately, temperature changes will greatly impact every  aspects of lake resource management.
+For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].
 During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].
-As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].
+As lake temperature increases (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].
 Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].
 Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.
-Ultimately, temperature changes will greatly impact every aspect of lake ecology and lake resource management.
+This need for accurate lake temperature forecasting will be crucial for protecting human and environmental health.
 
@@
 Models typically vary in number of lakes studied, complexity of modelling approach, and study interval.
-Modelling efforts include efforts to model a single lake over relatively small (e.g.
-hourly) time intervals [@saeed2016water; @peeters2002modeling; @zhong2016recent].
-Additionally, there are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
-In these efforts, air temperature and lake size are often the only selected predictor variables.
-There have also been some efforts aimed at modelling and measuring lake temperature across large spatial extents for large number of lakes [@minns2017factors; @o2015rapid; @wan2017comprehensive].
-These past efforts use a wide variety of approaches and have been rather successful at modelling lake photic zone temperature.
-One limitation, however, is that the majority of these lakes are large and few studies document or model smaller lakes.
-Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these smaller waterbodies [@downing2006global; @winslow2014lake].
+Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).
+Such studies include: [@saeed2016water; @peeters2002modeling; @zhong2016recent].
+There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].
+By and large, air temperature and lake size are often the only selected predictor variables.
+On the whole, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.
+Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].
+Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.
+Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes [@downing2006global; @winslow2014lake].
 
-In spite of the need to understand near-surface temperature for all lakes across a large spatial extent, there have been several challenges that have slowed our progress.
+Yet we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.
 Modelling lake temperature requires a large amount of data, and therefore study lakes are often selected opportunistically, which may introduce a spatial bias.
@@
 
-Our modelling effort takes advantage of the relatively recent availability of broad scale field data for lakes and uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
+The presented modelling effort uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
 The NLA is a stratified random sample of all lakes in the conterminous United States repeated every five years beginning in 2007.
@@
 
-The main goals of this work are to 1) develop a simple yet robust lake photic zone temperature model and 2) develop a model that is applicable to all lakes in the conterminous US that capture key drivers of near-surface lake temperature.
-Additionally, it is our practice conduct our work as openly as is feasible.
-Towards that end we provide access to all code and data used to develop these models with the active repository available at <https://github.com/usepa/lake_photic_zone> and the code, data, and documents are archived at <https://link_to_zenodo_right_before_submission>.
+To summarize, the main goal of this work was to develop a simple yet robust lake photic temperature model for all lakes in the conterminous US that capture key drivers of near-surface lake temperature.
+Additionally, this work (something about open science and GitHub link).
 
@@
 
-## Data
-
-We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment [@usepa2016national; @usepa2009national].
+##Data
+We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment (NLA) (CITE).
 The NLA is a stratified random sample of lakes (great than 1 ha) across the United States.
@@
 This research effort used the 2007 and 2012 sample years.
-The 2017 data are currently undergoing quality control before being released the public.
+The 2017 data are currently undergoing quality control before being released to the public.
 Upon release, the data can easily be included to improve this model.
-For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (`r fig_num("LakePhoticZonev2", caption = "Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes", display = "cite")`).
+For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).
 For each sampled lake, we used the mean temperature for all sampled depths of less than 2m (this being the depth typically considered the photic zone).
 
-![](../figures/LakePhoticZonev2.jpg)
-
-`r fig_num("LakePhoticZonev2")`
+![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
+Figure 1: Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes
 

### 2020-10-08 manuscript/manuscript.Rmd

--- 2020-10-02 manuscript/manuscript.Rmd
@@
 In addition to climate, there are additional factors clearly impacting lake temperature (e.g., surrounding land use, lake depth, size, and configuration, and elevation).
-To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages `lakemorpho` to calculate a suite of lake morphometry metric and `elevatr` to access digital elevation models for each lake [@hollister2017lakemorpho; @hollister2017elevatr].
+To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages `lakemorpho` to calculate a suite of lake morphometry metrics and `elevatr` to access digital elevation models for each lake [@hollister2017lakemorpho; @hollister2017elevatr].
 Additionally, the National Land Cover Database (NLCD) was our source for land cover data.
@@
 
-Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used to aid in variable selection and to calculate relative variable importance.
+Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and to calculate relative variable importance.
 Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
-Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.
-All random forest modelling was conducted in R, version 4.0.2 with the randomForest package [@R2020; @liaw2002rf].
-Model performance is reported as mean square error and adjusted R^2^.
+Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
+Model performance is reported as mean square error and adjusted R.
 

### 2020-11-05 manuscript/manuscript.Rmd
Added nla map, pred vs obs plot, and error map

--- 2020-10-02 manuscript/manuscript.Rmd
@@
 
-![`r fig_num("LakePhoticZonev2")`](../figures/LakePhoticZonev2.jpg)
+![`r fig_num("LakePhoticZonev2")`](../figures/nla_map.jpg)
 
@@
 
-In addition to a measure of overall model performance, we used percent increase of mean-squared error to assess variable importance.
+We evaluated the resultant model in several ways.
+First, we assess the overall model performance with traditional measures such as root mean square error of the out-of-bag residuals.
+Second, we examine error by comparing the predicted versus observed temperature for all lakes.
+In addition to these measures of overall model performance, we used percent increase of mean-squared error to assess variable importance.
 The percent increase in mean-squared error is a comparison between the true values of a variable and randomly permuted values of a specific variable on overall model performance.
@@
 less than 15 °C and greater than 32 °C) the model does not perform as well but lakes with growing season temperatures at these extremes were rare and represented only `r round(extreme$perc, 2)` of lakes sampled in the 2007 and 2012 NLA.
+Additionally, there is not an obvious spatial clustering of lakes with higher error (`r fig_num("obs_v_pred_map", caption = "Map of predicted minus observed temperature for 2007 and 2012 National Lakes Assessment lakes", display = "cite")`).
 
 ![`r fig_num("obs_v_pred")`](../figures/obs_v_pred.jpg)
+
+![`r fig_num("obs_v_pred_map")`](../figures/error_map.jpg)
 

### 2020-11-05 R/model.R

@@ -34,6 +34,7 @@ var_sel_tmean <- varsel_regression_rf(nla.all[,-1]$tmean_2m, select(nla.all[,-1]
 
 nla_select <- select(nla.all, nla_id, tmean_2m,var_sel_tmean$vars[[7]])
 write_csv(nla_select, here::here("data/nla_select.csv"))
+#nla_select <- read_csv(here::here("data/nla_select.csv"))
 set.seed(42)
 RFAll <- randomForest(tmean_2m ~ ., data = nla_select[,-1], importance = TRUE,
                ntree = 10000, keep.inbag = TRUE)
@@ -59,10 +60,20 @@ rf_all_trees <- data.frame(rf_all_trees)
 rf_all_trees <- mutate(rf_all_trees, nla_id = nla_select$nla_id, 
                        tmean_2m = nla_select$tmean_2m)
 rf_all_trees <- select(rf_all_trees, nla_id, tmean_2m, everything())
-rf_all_trees <- mutate(rf_all_trees, rmse = apply(rf_all_trees[,-1], 1, function(x) 
-  sqrt(mean((x[2:length(x)] - x[1])^2, na.rm = TRUE))))
-rf_all_rmse <- select(rf_all_trees, nla_id, tmean_2m, rmse) %>%
+rf_all_trees <- mutate(rf_all_trees, 
+                       mean_pred = apply(rf_all_trees[,-1], 1, 
+                                         function(x) mean(x[2:length(x)], 
+                                                               na.rm = TRUE)),
+                       rmse = apply(rf_all_trees[,-1], 1, 
+                                    function(x) sqrt(mean((x[2:length(x)] - 
+                                                             x[1])^2, 
+                                                          na.rm = TRUE))),
+                       mdev = apply(rf_all_trees[,-1], 1, 
+                                    function(x) mean(x[2:length(x)] - 
+                                                       x[1], na.rm = TRUE)))
+rf_all_rmse <- select(rf_all_trees, nla_id, tmean_2m, rmse, mdev, mean_pred) %>%
   left_join(nla)
+write_csv(rf_all_rmse, here::here("data/rf_all_tree_rmse.csv"))
 rmse_dens <- density(rf_all_rmse$rmse)
 #plot(rmse_dens)
 peak <- rmse_dens$x[which.max(rmse_dens$y)]

### 2020-12-22 manuscript/manuscript.Rmd
final edits to MS before internal review

@@
   Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
-- ^*2*^ ORISE, Narragansett, RI 02882
-- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
+
+- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*1*^, Hollister. J. W. ^*1*^, Milstead,
   W. Bryan. ^*1*^'
@@
 Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
-  | **Keywords:** Random forest, National Lakes Assessment, Limnology, water temperature, warming lakes
+  | **Keywords:** random forest, National Lakes Assessment, limnology, water temperature, warming lakes
 ---
@@
 One limitation, however, is that the majority of these lakes are large and few studies document or model smaller lakes.
-Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these smaller waterbodies [@downing2006global; @winslow2014lake].
+Considering that the number and importance of small lakes (< 1km^2^) has been historically underestimated, it is important to also predict temperatures in these smaller waterbodies [@downing2006global; @winslow2014lake].
 
@@
 When model efforts do attempt to cross large spatial extents, these efforts often rely on satellite data.
-While these models predict over large areas, they are restricted by the size of the lake captured by satellite (typically 3km2 for 1 km MODIS pixels).
+While these models predict over large areas, they are restricted by the size of the lake captured by satellite (typically 3km^2^ for 1 km MODIS pixels).
 
-Our modelling effort takes advantage of the relatively recent availability of broad scale field data for lakes and uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).
+Our modelling effort takes advantage of the relatively recent availability of broad scale field data for lakes and uses the US Environmental Protection Agency’s (EPA’s) National Lakes Assessment (NLA).
 The NLA is a stratified random sample of all lakes in the conterminous United States repeated every five years beginning in 2007.
@@
 at a large spatial extent), we included lakes with different morphologies, in different climates with diverse geologies, and surrounding landscapes.
-Additionally, small lakes were well represented with more than 50% of the lakes in the 2012 survey less than 0.5 km2.
+Additionally, small lakes were well represented with more than 50% of the lakes in the 2012 survey less than 0.5 km^2^.
 
@@
 Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).
-Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).
-Model performance is reported as mean square error and adjusted R.
+Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.All random forest modelling was conducted in R v 4.0.3 [@rcite] with the randomForest package [@liaw2020].
+Model performance is reported as mean square error and adjusted R^2^.
 
@@
 
-We used `r summ[summ$year == "2007",]$n` data points from the 2007 NLA and `r summ[summ$year == "2012",]$n` from 2012 NLA and these points are a spatially representative sample of lakes in the United States (`r fig_num("LakePhoticZonev2", display = "cite")`)
+We used `r summ[summ$year == "2007",]$n` data points from the 2007 NLA and `r summ[summ$year == "2012",]$n` from 2012 NLA and these points are a spatially representative sample of lakes in the United States (`r fig_num("LakePhoticZonev2", display = "cite")`).
 
@@
 The final model has a root mean square error of `r round(sqrt(mean(RFAll$mse)), 2)` and an adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.
-The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
+Despite general overall well-behaved performance, the final model fails to accurately predict lakes at extreme temperatures (`r fig_num("obs_v_pred", caption = "Predicted versus measured photic zone temperature", display = "cite")`).
+As might be expected, the model under estimates temperature for relatively high temperature lakes and over estimates colder lakes.
+Given the nature of random forest and the rarity of these events, it is difficult to improve the fit of extreme edges of the temperature distribution.
+It is expected that the addition of new NLA data will improve both the overall model and these rare events.
+Yet Figure 4 illustrates there is no spatial clustering to these error.The sampling date, the sampling day’s average ambient air temperature (data obtained from the PRISM Climate Group), and longitude are the most important variables impacting the final model’s accuracy.
 Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers.
@@
 Using the NLA data and simple metrics allowed this model to be applicable to any lake within the conterminous United States using minimal input data with a comparable RMSE.
-
-The final model included, average ambient air temperature and the average temperature of the prior 30 days.
-Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.
-It is likely that the three and seven day averages are not providing the model with unique information as the three and seven day averages are likely to be similar to the day-of average.
-Ambient air temperature does not change drastically over such a short time scale [CITATION??].
-Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.<!--JWH: We can quantify this with the data that we have.
-Should probably do this.
-Less arm wavey --> Yet the 30-day average does have an impact on lake photic zone temperature as it provides information about the longer term temperature intensity leading up to the sample date.
-Not just how far along into summer; that information would be captured in the date parameter.
-But more specifically, this measures the long term thermal heating happening at a site.
-
-# Alternate to the above paragraph
 
@@
 Sharma et al [-@sharma2015global] have compiled summer lake temperature data for 291 lakes for the period 1985-2009.
-This may be the largest lake temperature database to date; however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.
-<!--These data are open and published - accessible through the "laketemps" package.-->  One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.
+This may be the largest lake temperature database to date.
+One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.
 The model we present has proven to be accurate and will allow us to backcast lake temperatures for all the > 300,000 lakes included in NHDplus for the period of time covered by the PRISM climate predictions (1981 to present).

### 2021-08-26 manuscript/manuscript.Rmd
Adding edits to MS based on Frontiers review

@@
 ---
-title: "**Title:** Lake Photic Zone Temperature across the Conterminous United States"
+title: "**Title:** Predictive Model of Lake Photic Zone Temperature across the Conterminous United States"
 author:
-- ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
-  Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
+-'**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*1*^, Hollister. J. W. ^*1*^, Milstead, W.B.'^\*^ 
+- ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
 
-- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*1*^, Hollister. J. W. ^*1*^, Milstead,
-  W. Bryan. ^*1*^'
-- '^\*^  *corresponding author: kreakie.betty@epa.gov*'
+'**corresponding author: kreakie.betty@epa.gov*'
 bibliography: manuscript.bib
 output:
+  pdf_document: default
   word_document:
     reference_docx: manuscript_template_submission.docx
-  pdf_document: default
 csl: ecology.csl
 abstract: |
-  | **Abstract:** As the average global air temperature on Earth increases, surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).
-The influence of this increased temperature is known to impact lentic ecosystems across local to broad scales.
-For example, warming lake temperature is linked to disruptions in trophic linkages, changes in thermal stratification, and cyanobacteria bloom dynamics.
-Thus understanding broad trends in lake temperature is important to understanding the changing ecology of lakes and the potential human health impacts of these changes.
-To help address this, we developed a simple yet robust model of lake photic zone temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for the conterminous United States.
-The final model has a root mean square error of 1.48 °C and an adjusted R^2^ of 0.88.
-The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy.
+  | **Abstract:** As the average global air temperature increases, lake surface temperatures are also increasing globally.
+The influence of this increased temperature is known to impact lake ecosystems across local to broad scales.
+Warming lake temperature is linked to disruptions in trophic linkages, changes in thermal stratification, and cyanobacteria bloom dynamics.
+Thus, comprehending broad trends in lake temperature is important to understanding the changing ecology of lakes and the potential human health impacts of these changes.
+To help address this, we developed a simple yet robust random forest model of lake photic zone temperature using the 2007 and 2012 United States Environmental Protection Agency’s National Lakes Assessment data for the conterminous United States.
+The final model has a root mean square error of 1.48 °C and an adjusted R^^2^^ of 0.88; the final model included 2,282 total samples.
+The sampling date, that day’s average ambient air temperature and longitude are the most important variables impacting the final model’s accuracy.
+The final model also included 30-day average temperature, elevation, latitude, lake area, and lake shoreline length.
 Given the importance of temperature to a lake ecosystem, this model can be a valuable tool for researchers and lake resource managers.
-Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
-  | **Keywords:** random forest, National Lakes Assessment, limnology, water temperature, warming lakes
+Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information..
+  | **Keywords:** random forest, National Lakes Assessment, limnology, water temperature, warming lakes, photic zone temperature
 ---
